---
title: "explore HWE - wecare only"
output: html_document
---

started: Alexey Larionov, 2016  
last updated: Alexey Larionov, 12Apr2017

# Summary

Calculates AFs and HWE for 480 cases (before excluding eigenvectors outliers)  

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
#/analysis/mtgroup_share
#/scratch/medgen
setwd("/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts/wecare_only")
source_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/source_data"
interim_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts/wecare_only/interim_data"
results_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts/wecare_only/results"

# Required libraries
library(ggplot2)
library(plotly)
source("f04_hw_AL.R")
source("f03_qqunif_plot.R")

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "r04_filter_cases_and_variants_wecare_only.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(genotypes.mx)
class(genotypes.mx)
genotypes.mx[1:5,1:5]

dim(phenotypes.df)
str(phenotypes.df)
phenotypes.df[1:5,1:5]

dim(variants.df)
colnames(variants.df)
variants.df[1:5,1:5]

dim(kgen.df)
colnames(kgen.df)
kgen.df[1:5,1:5]

dim(exac.df)
colnames(exac.df)
exac.df[1:5,1:5]

# Check consistency of colnames and rownames
sum(colnames(genotypes.mx) != rownames(phenotypes.df))

sum(rownames(genotypes.mx) != rownames(variants.df))
sum(rownames(genotypes.mx) != rownames(kgen.df))
sum(rownames(genotypes.mx) != rownames(exac.df))

```

# recalculate_total_AFs

Used for selecting common variants for eigenvectors computation.  
Could be recalculated later after exclusion of eigenvectors outliers.  

```{r recalculate_total_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Rename AF fields in the variants table
vars_colnames <- colnames(variants.df)
"ac_raw" -> vars_colnames[ vars_colnames == "AC" ]
"an_raw" -> vars_colnames[ vars_colnames == "AN" ]
"af_raw" -> vars_colnames[ vars_colnames == "AF" ]
vars_colnames -> colnames(variants.df)

# Calculate total wecare ac, an and af
ac_all <- apply(genotypes.mx, 1, sum, na.rm=TRUE)
an_all <- apply(genotypes.mx, 1, get_allele_number.udf)
af_all <- ac_all/an_all

# Add new AFs to the variants table
variants.df <- cbind(variants.df, ac_all, an_all, af_all)

# Clean-up
rm(get_allele_number.udf, ac_all, an_all, af_all, vars_colnames)

```

# calculate_ubc_AFs

```{r calculate_ubc_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
ubc_cases <- phenotypes.df$cc == 0 
sum(ubc_cases) # 245
ubc_genotypes.mx <- genotypes.mx[,ubc_cases]
dim(ubc_genotypes.mx)

# Calculate ubc ac, an and af
ac_ubc <- apply(ubc_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc <- apply(ubc_genotypes.mx, 1, get_allele_number.udf)
af_ubc <- ac_ubc/an_ubc

# Add updated AFs to variants.df
variants.df <- cbind(variants.df, ac_ubc, an_ubc, af_ubc)

# Clean-up
rm(get_allele_number.udf, ubc_cases, ubc_genotypes.mx, ac_ubc, an_ubc, af_ubc)

```

# calculate_cbc_AFs

```{r calculate_cbc_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
cbc_cases <- phenotypes.df$cc == 1 
sum(cbc_cases) # 235
cbc_genotypes.mx <- genotypes.mx[,cbc_cases]
dim(cbc_genotypes.mx)

# Calculate cbc ac, an and af
ac_cbc <- apply(cbc_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc <- apply(cbc_genotypes.mx, 1, get_allele_number.udf)
af_cbc <- ac_cbc/an_cbc

# Add updated AFs to variants.df
variants.df <- cbind(variants.df, ac_cbc, an_cbc, af_cbc)

# Clean-up
rm(get_allele_number.udf, cbc_cases, cbc_genotypes.mx, ac_cbc, an_cbc, af_cbc)

```

# explore_hwe

echo=FALSE, message=FALSE, warning=FALSE  

```{r explore_hwe, results='hide', echo=FALSE, message=FALSE, warning=FALSE}

library(HardyWeinberg)

x <- MakeCounts(t(genotypes.mx),coding=c(0,1,2))
dim(x)
x[1:25,]

y <- HWExactStats(x[,1:3], verbose=FALSE)

# Select common variants
common_variants <- variants.df$af_all > 0.05 & variants.df$af_all < 0.95
sum(common_variants) # 48,574

# Bonferroni threshold - too relaxed (EZ)
th <- 1/length(y)
th # ~4e-6
hwe_violators <- y < th
sum(hwe_violators) # 724

th <- 0.00005 # Arbitrary convention between 5e-4 5e-5 (EZ)
hwe_violators <- y < th
sum(hwe_violators) # 801
qqunif.plot(y[!hwe_violators & common_variants], 
            main=paste("Common variants, excluding HWE violaters",
                       "\nexclusion threthold:", th, 
                       "\n excluded",sum(hwe_violators), "of", 
                       format(length(y), big.mark=",")))

th <- 0.0001 # Arbitrary convention between 5e-4 5e-5 (EZ)
hwe_violators <- y < th
sum(hwe_violators) # 827
qqunif.plot(y[!hwe_violators & common_variants], 
            main=paste("Common variants, excluding HWE violaters",
                       "\nexclusion threthold:", th, 
                       "\n excluded",sum(hwe_violators), "of", 
                       format(length(y), big.mark=",")))

th <- 0.0005 # Arbitrary convention between 5e-4 5e-5 (EZ)
hwe_violators <- y < th
sum(hwe_violators) # 918
qqunif.plot(y[!hwe_violators & common_variants], 
            main=paste("Common variants, excluding HWE violaters",
                       "\nexclusion threthold:", th, 
                       "\n excluded",sum(hwe_violators), "of", 
                       format(length(y), big.mark=",")))

th <- 0.001 # Stricter than arbitrary convention between 5e-4 5e-5 (EZ)
hwe_violators <- y < th
sum(hwe_violators) # 983
qqunif.plot(y[!hwe_violators & common_variants], 
            main=paste("Common variants, excluding HWE violaters",
                       "\nexclusion threthold:", th, 
                       "\n excluded",sum(hwe_violators), "of", 
                       format(length(y), big.mark=",")))

th <- 0.005 # Stricter than arbitrary convention between 5e-4 5e-5 (EZ)
hwe_violators <- y < th
sum(hwe_violators) # 1,954
qqunif.plot(y[!hwe_violators & common_variants], 
            main=paste("Common variants, excluding HWE violaters",
                       "\nexclusion threthold:", th, 
                       "\n excluded",format(sum(hwe_violators), big.mark=","), "of", 
                       format(length(y), big.mark=",")))

th <- 0.0001 # Arbitrary between 5e-4 5e-5, EZ
hwe_violators <- y < th
sum(hwe_violators)

z <- x[!hwe_violators & common_variants, 1:3]

HWQqplot(z, logplot=TRUE, pvaluetype="selome", 
         main="Q-Q Plot for HWE (w/o violators, common)")

# Plot 1/y against variants locations? 
# it seems that there could be spacial clusters of low p-values

```

# end_time

```{r end_time}

# Time stamp
Sys.time()

```

