---
title: "calculate_eigenvectors_wecare_nfe_kgen50_Mar2017"
output: html_document
---

started: Alexey Larionov, 2016  
last updated: Alexey Larionov, 30Mar2017

# Summary

Eigenvectors are calculated using common variants only:  
5% < AF in the dataset < 95%

Overall, eigenvectors are calculated for 3 datasets:  

- wecare-only
- wecare-nfe (using overlap wecare and nfe)
- wecare-nfe-50kgen (using overlap wecare, nfe and kgen50)  

This script deals with wecare-nfe-50kgen dataset 

Requires f01_calculate_eigenvectors.R

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
#/analysis/mtgroup_share
#/scratch /medgen
setwd("/analysis/mtgroup_share/scripts/wecare_stat_03.17/scripts")
source_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_03.17/source_data"
interim_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_03.17/interim_data"

# Required libraries
library(ggplot2)
library(plotly)

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "r04b_add_kgen50_Mar2017.RData", sep="/"))

```

# check_data

728 samples = 480 wecare + 198 nfe + 50 kgen50  

```{r check_data}

ls()

dim(genotypes.mx)
class(genotypes.mx)
genotypes.mx[1:5, 1:5]

dim(phenotypes.df)
str(phenotypes.df)
phenotypes.df[1:5, 1:5]

dim(variants.df)
str(variants.df)
variants.df[1:5,1:5]

dim(kgen.df)
str(kgen.df)
kgen.df[1:5,1:5]

dim(exac.df)
str(exac.df)
exac.df[1:5,1:5]

# Check consistence of rownames in kgen50_wecare_nfe_kgen50_genotypes.mx and kgen50_wecare_nfe_kgen50_variants.df
sum(rownames(genotypes.mx) != rownames(variants.df))
sum(rownames(genotypes.mx) != rownames(kgen.df))
sum(rownames(genotypes.mx) != rownames(exac.df))

sum(colnames(genotypes.mx) != rownames(phenotypes.df))

```

# make_wecare_nfe_kgen50_subset_of_data

keep 80,963 variants with kgen50 data

```{r make_wecare_nfe_kgen50_subset_of_data}

# variants with kgen50 data
kgen50_vars <- variants.df$kgen50
sum(kgen50_vars)

# Remove variants w/o kgen50 data
wecare_nfe_kgen50_variants.df <- variants.df[kgen50_vars,]
dim(wecare_nfe_kgen50_variants.df)

wecare_nfe_kgen50_genotypes.mx <- genotypes.mx[kgen50_vars,]
dim(wecare_nfe_kgen50_genotypes.mx)

wecare_nfe_kgen50_kgen.df <- kgen.df[kgen50_vars,]
dim(wecare_nfe_kgen50_kgen.df)

wecare_nfe_kgen50_exac.df <- exac.df[kgen50_vars,]
dim(wecare_nfe_kgen50_exac.df)

# rename unchaged data
wecare_nfe_kgen50_phenotypes.df <- phenotypes.df

# Clean-up
rm(kgen50_vars, genotypes.mx, phenotypes.df, variants.df, exac.df, kgen.df)

```

# check_for_variants_with_uniform_genotypes_accross_all_wecare_nfe_samples

There should be no such variants because they were removed earlier,  
when generated wecare-nfe and kgen50 datasets

```{r check_for_variants_with_uniform_genotypes_accross_all_wecare_nfe_samples}

# Check that there is no all-NA variants 
# (this is assumed for some downstream calculations)
non_NA_count.udf <- function(x){sum(!is.na(x))}
all_NA <- apply(wecare_nfe_kgen50_genotypes.mx, 1, non_NA_count.udf) == 0
sum(all_NA) # 0

# Function to detect uniform numeric vector
uniform_vector.udf <- function(x){
  if(min(x, na.rm=TRUE) == max(x, na.rm=TRUE)){return(TRUE)} else {return(FALSE)}}

# Variants with uniform genotypes accross all samples 
uniform_genotypes <- apply(wecare_nfe_kgen50_genotypes.mx, 1, uniform_vector.udf)
summary(uniform_genotypes)
sum(uniform_genotypes) # 0

# Clean-up
rm(non_NA_count.udf, all_NA, uniform_vector.udf, uniform_genotypes)

```

# recalculate_total_AFs

Used for selecting variants for eigenvectors computation.  
Could be recalculated later after exclusion of eigenvectors outliers.

```{r recalculate_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# --- Calculate total AFs --- #

# Rename AF fields in the variants table
vars_colnames <- colnames(wecare_nfe_kgen50_variants.df)
"ac_raw" -> vars_colnames[ vars_colnames == "AC" ]
"an_raw" -> vars_colnames[ vars_colnames == "AN" ]
"af_raw" -> vars_colnames[ vars_colnames == "AF" ]
vars_colnames -> colnames(wecare_nfe_kgen50_variants.df)

# Calculate new AFs
ac_all <- apply(wecare_nfe_kgen50_genotypes.mx, 1, sum, na.rm=TRUE)
an_all <- apply(wecare_nfe_kgen50_genotypes.mx, 1, get_allele_number.udf)
af_all <- ac_all/an_all

# Ceck new AFs 
# (note that uniform variants were excluded)
ac_all[1:10]
an_all[1:10]
af_all[1:10]

min(ac_all)
min(an_all)
min(af_all)

max(ac_all)
max(an_all)
max(af_all)

# Add new AFs to the variants table
wecare_nfe_kgen50_variants.df <- cbind(wecare_nfe_kgen50_variants.df, ac_all, an_all, af_all)

# Clean-up()
rm(get_allele_number.udf, vars_colnames, ac_all, an_all, af_all)

```

# calculate_kgen50_AFs

```{r calculate_kgen50_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
kgen50_cases <- wecare_nfe_kgen50_phenotypes.df$cc == -2
sum(kgen50_cases) # 50
kgen50_genotypes.mx <- wecare_nfe_kgen50_genotypes.mx[,kgen50_cases]
dim(kgen50_genotypes.mx)

# Calculate ubc ac, an and af
ac_kgen50 <- apply(kgen50_genotypes.mx, 1, sum, na.rm=TRUE)
an_kgen50 <- apply(kgen50_genotypes.mx, 1, get_allele_number.udf)
af_kgen50 <- ac_kgen50/an_kgen50

# Add updated AFs to wecare_variants.df
wecare_nfe_kgen50_variants.df <- cbind(wecare_nfe_kgen50_variants.df, ac_kgen50, an_kgen50, af_kgen50)

# Clean-up
rm(get_allele_number.udf, kgen50_cases, kgen50_genotypes.mx, ac_kgen50, an_kgen50, af_kgen50)

```

# calculate_nfe_AFs

```{r calculate_nfe_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
nfe_cases <- wecare_nfe_kgen50_phenotypes.df$cc == -1
sum(nfe_cases) # 198
nfe_genotypes.mx <- wecare_nfe_kgen50_genotypes.mx[,nfe_cases]
dim(nfe_genotypes.mx)

# Calculate ubc ac, an and af
ac_nfe <- apply(nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_nfe <- apply(nfe_genotypes.mx, 1, get_allele_number.udf)
af_nfe <- ac_nfe/an_nfe

# Add updated AFs to wecare_variants.df
wecare_nfe_kgen50_variants.df <- cbind(wecare_nfe_kgen50_variants.df, ac_nfe, an_nfe, af_nfe)

# Clean-up
rm(get_allele_number.udf, nfe_cases, nfe_genotypes.mx, ac_nfe, an_nfe, af_nfe)

```

# calculate_wecare_AFs

```{r calculate_wecare_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
wecare_cases <- wecare_nfe_kgen50_phenotypes.df$cc > -1
sum(wecare_cases) # 198
wecare_genotypes.mx <- wecare_nfe_kgen50_genotypes.mx[,wecare_cases]
dim(wecare_genotypes.mx)

# Calculate ubc ac, an and af
ac_wecare <- apply(wecare_genotypes.mx, 1, sum, na.rm=TRUE)
an_wecare <- apply(wecare_genotypes.mx, 1, get_allele_number.udf)
af_wecare <- ac_wecare/an_wecare

# Add updated AFs to wecare_variants.df
wecare_nfe_kgen50_variants.df <- cbind(wecare_nfe_kgen50_variants.df, ac_wecare, an_wecare, af_wecare)

# Clean-up
rm(get_allele_number.udf, wecare_cases, wecare_genotypes.mx, ac_wecare, an_wecare, af_wecare)

```

# calculate_eigenvectors

Only common variants will be used for eigenvectors calculation.  

Note exclusion on both sides: low- and high- AFs:  
- Low AFs remove rare variants with common allele in reference genome  
- Hight AFs remove rare variants with common allele in reference genome  

```{r calculate_eigenvectors}

# --- Make subset of data for eigenvectors calculation --- #

# Select kgen50_wecare-nfe overlap variants
kgen50_common <- wecare_nfe_kgen50_variants.df$af_kgen50 > 0.05 & wecare_nfe_kgen50_variants.df$af_kgen50 < 0.95
nfe_common <- wecare_nfe_kgen50_variants.df$af_nfe > 0.05 & wecare_nfe_kgen50_variants.df$af_nfe < 0.95
wecare_common <- wecare_nfe_kgen50_variants.df$af_wecare > 0.05 & wecare_nfe_kgen50_variants.df$af_wecare < 0.95

sum(kgen50_common) # 44,421
sum(nfe_common) # 42,843
sum(wecare_common) # 42,234

common_overlap_vars <- kgen50_common & nfe_common & wecare_common
sum(common_overlap_vars) # 36,539

# Select common variants
min(wecare_nfe_kgen50_variants.df$af_all[common_overlap_vars]) # ~0.05
max(wecare_nfe_kgen50_variants.df$af_all[common_overlap_vars]) # ~0.95

# Make the subset of data
common_overlap_genotypes.mx <- wecare_nfe_kgen50_genotypes.mx[common_overlap_vars,]
dim(common_overlap_genotypes.mx)
common_overlap_genotypes.mx[1:5,1:5]

# --- Calculate eigenvectors --- #

source("f01_calculate_eigenvectors.R")
wecare_nfe_kgen50.eigen <- normalise_and_calculate_eigenvectors.udf(common_overlap_genotypes.mx)

# Clean-up
rm(wecare_common, nfe_common, kgen50_common, common_overlap_vars, 
   normalise_and_calculate_eigenvectors.udf, common_overlap_genotypes.mx)

```

# plot_eigenvectors

```{r plot_eigenvectors}

# --- Prepare data for plotting --- #

eigenvectors.mx <- as.data.frame(wecare_nfe_kgen50.eigen$vectors) # eigenvectors in columns

# Prepare cases IDs
cases_IDs <- colnames(wecare_nfe_kgen50_genotypes.mx)

# Prepare cases lables
kgen50_cases <- wecare_nfe_kgen50_phenotypes.df$cc == -2
kgen50_populations <- wecare_nfe_kgen50_phenotypes.df[kgen50_cases,"population"]
cases_labels <- c(kgen50_populations, rep("NFE",198), rep("WECARE",480))
cases_labels <- factor(cases_labels, levels=c("WECARE", "NFE", "AFR", "AMR", "SAS", "EAS", "EUR"))
# plotly layers dots in order of the labels' factor levels. 
# Thus I made EUR above NFE and WECARE
# By default the order would be sort(unique(..))

# make the dataframe
data2plot.df <- cbind(cases_IDs, cases_labels, eigenvectors.mx[,1:5])
colnames(data2plot.df) <- c("sample", "group", "ev1", "ev2", "ev3", "ev4", "ev5")

# Prepare colour scale
colours <- c("EUR" = "BLUE", "AFR" = "BROWN", "AMR" = "GREEN", "SAS" = "BLACK", "EAS" = "MAGENTA", "NFE" = "PINK", "WECARE" = "RED")
userColourScale <- scale_colour_manual(values=colours)

# --- Plot eig1 vs eig2 --- #

g <- ggplot(data2plot.df, aes(ev2, ev1)) +
  geom_point(aes(col=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>common overlapped variants (37,037 x 728)", 
       x = "eigenvector2", y ="eigenvector1") +
  userColourScale

ggplotly(g) # Coloured dots are layered in order of the group's factor's levels

# --- Plot eig2 vs eig3 --- #

g <- ggplot(data2plot.df, aes(ev2, ev3)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>common overlapped variants (37,037 x 728)", x ="eigenvector2", y = "eigenvector3") +
  userColourScale

ggplotly(g)

# --- Plot eig3 vs eig4 --- #

g <- ggplot(data2plot.df, aes(ev3, ev4)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>common overlapped variants (37,037 x 728)", x ="eigenvector3", y = "eigenvector4") +
  userColourScale

ggplotly(g)

# --- Plot eig4 vs eig5 --- #

g <- ggplot(data2plot.df, aes(ev4, ev5)) +
  geom_point(aes(colour=group, fill=group, text = cases_IDs)) + 
  labs(title="wecare-nfe-kgen50<br>common overlapped variants (37,037 x 728)", x ="eigenvector4", y = "eigenvector5") +
  userColourScale

ggplotly(g)

# --- Clean-up --- #

rm(eigenvectors.mx, colours, g, data2plot.df, kgen50_populations, 
   kgen50_cases, cases_IDs, cases_labels, userColourScale)

```

# calculate_wecare_outliers

Explore 6 standard deviations in 5 top eigenvectors  

```{r calculate_wecare_outliers}

eigenvectors.mx <- as.data.frame(wecare_nfe_kgen50.eigen$vectors) # eigenvectors in columns, samples in rows

wecare_samples <- c(rep(FALSE,50), rep(FALSE, 198), rep(TRUE, 480))
wecare_samples_ids <- colnames(wecare_nfe_kgen50_genotypes.mx)[wecare_samples]

wecare_eigenvectors.mx <- eigenvectors.mx[wecare_samples,]

ev1 <- wecare_eigenvectors.mx[,1]
ev1.positive_outliers <- ev1 > mean(ev1) + 6 * sd(ev1)
ev1.negative_outliers <- ev1 < mean(ev1) - 6 * sd(ev1)
sum(ev1.positive_outliers)
sum(ev1.negative_outliers)

wecare_samples_ids[ev1.positive_outliers]
wecare_samples_ids[ev1.negative_outliers] # P5_E09, P6_D05

ev2 <- wecare_eigenvectors.mx[,2]
ev2.positive_outliers <- ev2 > mean(ev2) + 6 * sd(ev2)
ev2.negative_outliers <- ev2 < mean(ev2) - 6 * sd(ev2)
sum(ev2.positive_outliers)
sum(ev2.negative_outliers)

wecare_samples_ids[ev2.positive_outliers]
wecare_samples_ids[ev2.negative_outliers] # P5_E09, P6_D05

ev3 <- wecare_eigenvectors.mx[,3]
ev3.positive_outliers <- ev3 > mean(ev3) + 6 * sd(ev3)
ev3.negative_outliers <- ev3 < mean(ev3) - 6 * sd(ev3)
sum(ev3.positive_outliers)
sum(ev3.negative_outliers)

wecare_samples_ids[ev3.positive_outliers]
wecare_samples_ids[ev3.negative_outliers]

ev4 <- wecare_eigenvectors.mx[,4]
ev4.positive_outliers <- ev4 > mean(ev4) + 6 * sd(ev4)
ev4.negative_outliers <- ev4 < mean(ev4) - 6 * sd(ev4)
sum(ev4.positive_outliers)
sum(ev4.negative_outliers)

wecare_samples_ids[ev4.positive_outliers]
wecare_samples_ids[ev4.negative_outliers]

ev5 <- wecare_eigenvectors.mx[,5]
ev5.positive_outliers <- ev5 > mean(ev5) + 6 * sd(ev5)
ev5.negative_outliers <- ev5 < mean(ev5) - 6 * sd(ev5)
sum(ev5.positive_outliers)
sum(ev5.negative_outliers)

wecare_samples_ids[ev5.positive_outliers]
wecare_samples_ids[ev5.negative_outliers]

# Clean-up
rm(wecare_samples, wecare_samples_ids, 
   eigenvectors.mx, wecare_eigenvectors.mx, 
   ev1, ev1.positive_outliers, ev1.negative_outliers, 
   ev2, ev2.positive_outliers, ev2.negative_outliers, 
   ev3, ev3.positive_outliers, ev3.negative_outliers,
   ev4, ev4.positive_outliers, ev4.negative_outliers,
   ev5, ev5.positive_outliers, ev5.negative_outliers)

```

# data_summary

```{r data_summary}

ls()

dim(wecare_nfe_kgen50_genotypes.mx)
class(wecare_nfe_kgen50_genotypes.mx)
wecare_nfe_kgen50_genotypes.mx[1:5,1:5]

dim(wecare_nfe_kgen50_variants.df)
str(wecare_nfe_kgen50_variants.df)
wecare_nfe_kgen50_variants.df[1:5,1:5]

dim(wecare_nfe_kgen50_phenotypes.df)
colnames(wecare_nfe_kgen50_phenotypes.df)
wecare_nfe_kgen50_phenotypes.df[1:5,]

dim(wecare_nfe_kgen50_kgen.df)
colnames(wecare_nfe_kgen50_kgen.df)
wecare_nfe_kgen50_kgen.df[1:5,]

dim(wecare_nfe_kgen50_exac.df)
colnames(wecare_nfe_kgen50_exac.df)
wecare_nfe_kgen50_exac.df[1:5,]

str(wecare_nfe_kgen50.eigen)

sum(rownames(wecare_nfe_kgen50_genotypes.mx) != rownames(wecare_nfe_kgen50_variants.df))
sum(rownames(wecare_nfe_kgen50_genotypes.mx) != rownames(wecare_nfe_kgen50_kgen.df))
sum(rownames(wecare_nfe_kgen50_genotypes.mx) != rownames(wecare_nfe_kgen50_exac.df))
sum(colnames(wecare_nfe_kgen50_genotypes.mx) != rownames(wecare_nfe_kgen50_phenotypes.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r05c_calculate_egenvectors_wecare_nfe_kgen50_Mar2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
