---
title: "pick_eigenvectors_exclude_outliers_Mar2017"
output: html_document
---

started: Alexey Larionov, 16Feb2017  
last updated: Alexey Larionov, 30Mar2017

# Summary

- Add eigenvectors to wecare-nfe data  

- Exclude ethinic outliers: P5_E09 and P6_D05  
  (detected with kgen50 eigenvectors, agreed with DC on Skype 28Mar2017)  

- Export the final phenotype table  

- Recalculate allelic frequencies  

Remaining dataset contains:  
  676 cases x 274,037 variants  
  478 WECARE  
  198 NFE  
  233 CBC  
  245 UBC  
  84 CBC-Fam  
  150 CBC-noFam  
  71 UBC-Fam  
  174 UBC-noFam  
  225 matched CBC-UBC pairs  

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
#/analysis/mtgroup_share
#/scratch /medgen
setwd("/analysis/mtgroup_share/scripts/wecare_stat_03.17/scripts")
source_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_03.17/source_data"
interim_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_03.17/interim_data"
results_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_03.17/results"

```

# load_wecare_only_eigenvectors

```{r read_wecare_only_eigenvectors}

load(paste(interim_data_folder, "r05a_calculate_egenvectors_wecare_only_Mar2017.RData", sep="/"))
wecare_phenotypes.df$wes_id -> rownames(wecare.eigen$vectors) 
wecare.eigen$vectors[1:5,1:5]
rm(wecare_genotypes.mx, wecare_phenotypes.df, wecare_variants.df, wecare_kgen.df, wecare_exac.df)

```

# load_wecare_nfe_eigenvectors

```{r load_wecare_nfe_eigenvectors}

load(paste(interim_data_folder, "r05b_calculate_egenvectors_wecare_nfe_Mar2017.RData", sep="/"))
wecare_nfe_phenotypes.df$wes_id -> rownames(wecare_nfe.eigen$vectors) 
wecare_nfe.eigen$vectors[1:5,1:5]
rm(wecare_nfe_genotypes.mx, wecare_nfe_phenotypes.df, wecare_nfe_variants.df, 
   wecare_nfe_kgen.df, wecare_nfe_exac.df)

```

# load_wecare_nfe_kgen50_eigenvectors

```{r load_wecare_nfe_kgen50_eigenvectors}

load(paste(interim_data_folder, "r05c_calculate_egenvectors_wecare_nfe_kgen50_Mar2017.RData", sep="/"))
wecare_nfe_kgen50_phenotypes.df$wes_id -> rownames(wecare_nfe_kgen50.eigen$vectors) 
wecare_nfe_kgen50.eigen$vectors[1:5,1:5]
rm(wecare_nfe_kgen50_genotypes.mx, wecare_nfe_kgen50_phenotypes.df, wecare_nfe_kgen50_variants.df, 
   wecare_nfe_kgen50_kgen.df, wecare_nfe_kgen50_exac.df)

```

# load_wecare_nfe_data

```{r load_wecare_nfe_data}

load(paste(interim_data_folder, "r04a_filter_cases_and_variants_Mar2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(wecare_nfe_variants.df)
colnames(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

# Check consistency of colnames and rownames
sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))

```

# add_kgen50_eigenvectors_to_phenotypes_table

```{r add_wes_eigenvectors_to_phenotypes_table}

# Prepare matrix with top 3 eigenvectors
wecare_nfe_cases <- wecare_nfe_phenotypes.df$wes_id
kgen50_eigenvectors.mx <- wecare_nfe_kgen50.eigen$vectors[wecare_nfe_cases,1:3]
dim(kgen50_eigenvectors.mx)
colnames(kgen50_eigenvectors.mx) <- c("eig1_kgen50", "eig2_kgen50", "eig3_kgen50")

# Add columns to phenotypes table
wecare_nfe_phenotypes.df <- cbind(wecare_nfe_phenotypes.df, kgen50_eigenvectors.mx)
str(wecare_nfe_phenotypes.df)

# Clean-up
rm(wecare_nfe_cases, kgen50_eigenvectors.mx, wecare_nfe_kgen50.eigen)

```

# add_wecare_nfe_eigenvectors_to_phenotypes_table

```{r add_wecare_nfe_eigenvectors_to_phenotypes_table}

# Prepare matrix with top 3 eigenvectors
wecare_nfe_cases <- wecare_nfe_phenotypes.df$wes_id
nfe_eigenvectors.mx <- wecare_nfe.eigen$vectors[wecare_nfe_cases,1:3]
dim(nfe_eigenvectors.mx)
colnames(nfe_eigenvectors.mx) <- c("eig1_nfe", "eig2_nfe", "eig3_nfe")

# Add columns to phenotypes table
wecare_nfe_phenotypes.df <- cbind(wecare_nfe_phenotypes.df, nfe_eigenvectors.mx)
str(wecare_nfe_phenotypes.df)

# Clean-up
rm(wecare_nfe_cases, nfe_eigenvectors.mx, wecare_nfe.eigen)

```

# add_wecare_only_eigenvectors_to_phenotypes_table

```{r add_wecare_only_eigenvectors_to_phenotypes_table}

# Prepare matrix with top 3 eigenvectors
wecare_cases <- wecare_nfe_phenotypes.df[wecare_nfe_phenotypes.df$cc > -1,"wes_id"]
wecare_eigenvectors.mx <- wecare.eigen$vectors[wecare_cases,1:3]
dim(wecare_eigenvectors.mx)
colnames(wecare_eigenvectors.mx) <- c("eig1_wecare", "eig2_wecare", "eig3_wecare")

# Add NA rows for nfe cases (to fit the wecare_nfe_phenotypes.df)
wecare_eigenvectors_2add.mx <- matrix(rep(NA,nrow(wecare_nfe_phenotypes.df)*3), ncol=3)
colnames(wecare_eigenvectors_2add.mx) <- c("eig1_wecare", "eig2_wecare", "eig3_wecare")
rownames(wecare_eigenvectors_2add.mx) <- wecare_nfe_phenotypes.df$wes_id

for(id in rownames(wecare_eigenvectors.mx)){
  wecare_eigenvectors_2add.mx[id,] <- wecare_eigenvectors.mx[id,]
}

# Add columns to phenotypes table
wecare_nfe_phenotypes.df <- cbind(wecare_nfe_phenotypes.df, wecare_eigenvectors_2add.mx)
str(wecare_nfe_phenotypes.df)

# Clean-up
rm(wecare_cases, wecare_eigenvectors.mx, wecare_eigenvectors_2add.mx, id, wecare.eigen)

```

# make_final_phenotypes_file

Add new eigenvectors and flag eigenvectors outliers  
(P6_D05, P5_E09 - selected earlier by kgen50 eigenvectors  
Agreed with DC by Skype on 28Mar2017)  

```{r make_final_phenotypes_file}

# List of eigenvectors outliers
eig_outliers=c("P6_D05", "P5_E09")

# Read previous phenotype file (w/o new outliers and eigenvectors)
phenotypes.df <- read.table(paste(interim_data_folder, "wecare_phenotypes.txt", sep="/"), 
                            header = TRUE, quote="", sep="\t")

phenotypes.df[1:5,1:5] # apparently it already has rownames

# Subset wecare-nfe phenotype data frame (w/o excluded cases)
wecare_cases <- wecare_nfe_phenotypes.df$cc != -1
updated_phenotypes.df <- wecare_nfe_phenotypes.df[wecare_cases,]
dim(updated_phenotypes.df)
updated_phenotypes.df[1:5,1:5]

# Prepare matrix with eigenvectors to add
cols_2add <- c("eig1_kgen50", "eig2_kgen50", "eig3_kgen50", 
               "eig1_nfe", "eig2_nfe", "eig3_nfe", 
               "eig1_wecare", "eig2_wecare", "eig3_wecare")
eigs_2add.mx <- matrix(as.numeric(rep(NA,nrow(phenotypes.df)*9)), ncol=9)
colnames(eigs_2add.mx) <- cols_2add
rownames(eigs_2add.mx) <- rownames(phenotypes.df)

wecare_nfe_phenotypes_eigs.mx <- as.matrix(wecare_nfe_phenotypes.df[, cols_2add])

overlapped_cases <- intersect(rownames(wecare_nfe_phenotypes_eigs.mx), rownames(eigs_2add.mx))
length(overlapped_cases) # 480

for(id in overlapped_cases){
  eigs_2add.mx[id,] <- wecare_nfe_phenotypes_eigs.mx[id,]
}

# Add the new eigenvectors to the phenotype file
phenotypes.df <- cbind(phenotypes.df, eigs_2add.mx)

# Add eigenvectors outliers
phenotypes.df$filter <- as.character(as.vector(phenotypes.df$filter))
"eigenvector_outlier" -> phenotypes.df[eig_outliers,"filter"]

# Write down the updated phenotype file
write.table(phenotypes.df, file=paste(results_folder, "wecare_phenotypes.txt", sep="/"), quote=FALSE, sep="\t")

# Clean-up
rm(eig_outliers, phenotypes.df, wecare_cases, updated_phenotypes.df, id, 
   cols_2add, eigs_2add.mx, wecare_nfe_phenotypes_eigs.mx, overlapped_cases)

```

# remove_wes_eigenvectors_outliers

Remove 2 cases: 678 -> 676  

Excluding P6_D05, P5_E09  
Selected earlier by kgen50 eigenvectors  
Agreed with DC by Skype on 28Mar2017  

```{r remove_wes_eigenvectors_outliers}

# Flag outliers
eig_outliers=c("P6_D05", "P5_E09")
"eigenvector_outlier" -> 
  wecare_nfe_phenotypes.df[wecare_nfe_phenotypes.df$wes_id %in% eig_outliers,"filter"]

# Count outliers
outliers <- wecare_nfe_phenotypes.df$filter == "eigenvector_outlier"
sum(outliers)

# Remove outliers
wecare_nfe_phenotypes.df <- wecare_nfe_phenotypes.df[!outliers,]
wecare_nfe_genotypes.mx <- wecare_nfe_genotypes.mx[,!outliers]

dim(wecare_nfe_phenotypes.df)
dim(wecare_nfe_genotypes.mx)

# Clean-up
rm(eig_outliers, outliers)

```

# remove_variants_with_uniform_genotypes_accross_all_samples
Remove 1,479 variants: 275,516 -> 274,037

```{r remove_variants_with_uniform_genotypes_accross_all_samples}

# Check that there is no all-NA variants
non_NA_count.udf <- function(x){sum(!is.na(x))}
all_NA <- apply(wecare_nfe_genotypes.mx, 1, non_NA_count.udf) == 0
sum(all_NA) # 0

# Function to detect uniform numeric vector
uniform_vector.udf <- function(x){
  if(min(x, na.rm=TRUE) == max(x, na.rm=TRUE)){return(TRUE)} else {return(FALSE)}}

# Variants with uniform genotypes accross all samples 
uniform_genotypes <- apply(wecare_nfe_genotypes.mx, 1, uniform_vector.udf)
summary(uniform_genotypes)
sum(uniform_genotypes) # 1,479

# Remove variants with uniform genotypes accross all samples
wecare_nfe_genotypes.mx <- wecare_nfe_genotypes.mx[!uniform_genotypes,]
dim(wecare_nfe_genotypes.mx) # 274,037    676
wecare_nfe_genotypes.mx[1:5,1:5]

wecare_nfe_variants.df <- wecare_nfe_variants.df[!uniform_genotypes,]
dim(wecare_nfe_variants.df) # 274,037     23
wecare_nfe_variants.df[1:5,1:5]

wecare_nfe_kgen.df <- wecare_nfe_kgen.df[!uniform_genotypes,]
dim(wecare_nfe_kgen.df) # 274,037      9
wecare_nfe_kgen.df[1:5,1:5]

wecare_nfe_exac.df <- wecare_nfe_exac.df[!uniform_genotypes,]
dim(wecare_nfe_exac.df) # 274,037     48
wecare_nfe_exac.df[1:5,1:5]

# Clean-up
rm(non_NA_count.udf, all_NA, uniform_vector.udf, uniform_genotypes)

```

# explore_remaining_cases

233 CBC, 245 UBC, 198 NFE
225 matched pairs

```{r explore_remaining_cases}

# Numbers of cases and controls
summary(as.factor(wecare_nfe_phenotypes.df$cc))
# -1   0   1 
#198 245 233 

# Number of matched pairs
sum(table(wecare_nfe_phenotypes.df$setno)==2) # 225

```

# calculate_total_AFs

```{r calculate_total_AFs}

# Function to count alleles (2x for diploid)
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Rename AF fields in the variants table
vars_colnames <- colnames(wecare_nfe_variants.df)
"ac_raw" -> vars_colnames[ vars_colnames == "AC" ]
"an_raw" -> vars_colnames[ vars_colnames == "AN" ]
"af_raw" -> vars_colnames[ vars_colnames == "AF" ]
vars_colnames -> colnames(wecare_nfe_variants.df)

# Calculate new AFs
ac_all <- apply(wecare_nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_all <- apply(wecare_nfe_genotypes.mx, 1, get_allele_number.udf)
af_all <- ac_all/an_all

# Ceck new AFs 
# (note that uniform variants were excluded)
ac_all[1:10]
an_all[1:10]
af_all[1:10]

min(ac_all)
min(an_all)
min(af_all)

max(ac_all)
max(an_all)
max(af_all)

# Add new AFs to the variants table
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_all, an_all, af_all)

# Clean-up()
rm(vars_colnames, ac_all, an_all, af_all)

```

# calculate_nfe_AFs

```{r calculate_nfe_AFs}

# Prepare genotypes table
nfe_cases <- wecare_nfe_phenotypes.df$cc == -1 
sum(nfe_cases) # 198
nfe_genotypes.mx <- wecare_nfe_genotypes.mx[,nfe_cases]
dim(nfe_genotypes.mx)

# Calculate nfe ac, an and af
ac_nfe <- apply(nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_nfe <- apply(nfe_genotypes.mx, 1, get_allele_number.udf)
af_nfe <- ac_nfe/an_nfe

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_nfe, an_nfe, af_nfe)

# Clean-up
rm(nfe_cases, nfe_genotypes.mx, ac_nfe, an_nfe, af_nfe)

```

# calculate_wecare_AFs

```{r calculate_wecare_AFs}

# Prepare genotypes table
wecare_cases <- wecare_nfe_phenotypes.df$cc == 0 | wecare_nfe_phenotypes.df$cc == 1
sum(wecare_cases) # 478
wecare_genotypes.mx <- wecare_nfe_genotypes.mx[,wecare_cases]
dim(wecare_genotypes.mx)

# Calculate nfe ac, an and af
ac_wecare <- apply(wecare_genotypes.mx, 1, sum, na.rm=TRUE)
an_wecare <- apply(wecare_genotypes.mx, 1, get_allele_number.udf)
af_wecare <- ac_wecare/an_wecare

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_wecare, an_wecare, af_wecare)

# Clean-up
rm(wecare_cases, wecare_genotypes.mx, ac_wecare, an_wecare, af_wecare)

```

# calculate_ubc_AFs

```{r calculate_ubc_AFs}

# Prepare genotypes table
ubc_cases <- wecare_nfe_phenotypes.df$cc == 0 
sum(ubc_cases) # 245
ubc_genotypes.mx <- wecare_nfe_genotypes.mx[,ubc_cases]
dim(ubc_genotypes.mx)

# Calculate ubc ac, an and af
ac_ubc <- apply(ubc_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc <- apply(ubc_genotypes.mx, 1, get_allele_number.udf)
af_ubc <- ac_ubc/an_ubc

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_ubc, an_ubc, af_ubc)

# Clean-up
rm(ubc_cases, ubc_genotypes.mx, ac_ubc, an_ubc, af_ubc)

```

# calculate_cbc_AFs

```{r calculate_cbc_AFs}

# Prepare genotypes table
cbc_cases <- wecare_nfe_phenotypes.df$cc == 1 
sum(cbc_cases) # 233
cbc_genotypes.mx <- wecare_nfe_genotypes.mx[,cbc_cases]
dim(cbc_genotypes.mx)

# Calculate cbc ac, an and af
ac_cbc <- apply(cbc_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc <- apply(cbc_genotypes.mx, 1, get_allele_number.udf)
af_cbc <- ac_cbc/an_cbc

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_cbc, an_cbc, af_cbc)

# Clean-up
rm(cbc_cases, cbc_genotypes.mx, ac_cbc, an_cbc, af_cbc)

```

# calculate_ubc_fam_AFs

```{r calculate_ubc_fam_AFs}

# Prepare genotypes table
ubc_fam_cases <- wecare_nfe_phenotypes.df$cc == 0 & wecare_nfe_phenotypes.df$family_history == 1
sum(ubc_fam_cases) # 71
ubc_fam_genotypes.mx <- wecare_nfe_genotypes.mx[,ubc_fam_cases]
dim(ubc_fam_genotypes.mx)

# Calculate ubc_fam ac, an and af
ac_ubc_fam <- apply(ubc_fam_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc_fam <- apply(ubc_fam_genotypes.mx, 1, get_allele_number.udf)
af_ubc_fam <- ac_ubc_fam/an_ubc_fam

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_ubc_fam, an_ubc_fam, af_ubc_fam)

# Clean-up
rm(ubc_fam_cases, ubc_fam_genotypes.mx, ac_ubc_fam, an_ubc_fam, af_ubc_fam)

```

# calculate_ubc_nofam_AFs

```{r calculate_ubc_nofam_AFs}

# Prepare genotypes table
ubc_nofam_cases <- wecare_nfe_phenotypes.df$cc == 0 & wecare_nfe_phenotypes.df$family_history == 0
sum(ubc_nofam_cases) # 174
ubc_nofam_genotypes.mx <- wecare_nfe_genotypes.mx[,ubc_nofam_cases]
dim(ubc_nofam_genotypes.mx)

# Calculate ubc_nofam ac, an and af
ac_ubc_nofam <- apply(ubc_nofam_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc_nofam <- apply(ubc_nofam_genotypes.mx, 1, get_allele_number.udf)
af_ubc_nofam <- ac_ubc_nofam/an_ubc_nofam

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_ubc_nofam, an_ubc_nofam, af_ubc_nofam)

# Clean-up
rm(ubc_nofam_cases, ubc_nofam_genotypes.mx, ac_ubc_nofam, an_ubc_nofam, af_ubc_nofam)

```

# calculate_cbc_fam_AFs

```{r calculate_cbc_fam_AFs}

# Prepare genotypes table
cbc_fam_cases <- wecare_nfe_phenotypes.df$cc == 1 & wecare_nfe_phenotypes.df$family_history == 1
sum(cbc_fam_cases) # 83
cbc_fam_genotypes.mx <- wecare_nfe_genotypes.mx[,cbc_fam_cases]
dim(cbc_fam_genotypes.mx)

# Calculate cbc_fam ac, an and af
ac_cbc_fam <- apply(cbc_fam_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc_fam <- apply(cbc_fam_genotypes.mx, 1, get_allele_number.udf)
af_cbc_fam <- ac_cbc_fam/an_cbc_fam

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_cbc_fam, an_cbc_fam, af_cbc_fam)

# Clean-up
rm(cbc_fam_cases, cbc_fam_genotypes.mx, ac_cbc_fam, an_cbc_fam, af_cbc_fam)

```

# calculate_cbc_nofam_AFs

```{r calculate_cbc_nofam_AFs}

# Prepare genotypes table
cbc_nofam_cases <- wecare_nfe_phenotypes.df$cc == 1 & wecare_nfe_phenotypes.df$family_history == 0
sum(cbc_nofam_cases) # 150
cbc_nofam_genotypes.mx <- wecare_nfe_genotypes.mx[,cbc_nofam_cases]
dim(cbc_nofam_genotypes.mx)

# Calculate cbc_nofam ac, an and af
ac_cbc_nofam <- apply(cbc_nofam_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc_nofam <- apply(cbc_nofam_genotypes.mx, 1, get_allele_number.udf)
af_cbc_nofam <- ac_cbc_nofam/an_cbc_nofam

# Add updated AFs to wecare_nfe_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_cbc_nofam, an_cbc_nofam, af_cbc_nofam)

# Clean-up
rm(cbc_nofam_cases, cbc_nofam_genotypes.mx, ac_cbc_nofam, 
   an_cbc_nofam, af_cbc_nofam, get_allele_number.udf)

```

# data_summary

```{r data_summary}

ls()

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(wecare_nfe_variants.df)
colnames(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))

sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r06_pick_egenvectors_exclude_outliers_Mar2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
