---
title: "calculate_eigenvectors_wecare_nfe_Apr2017"
output: html_document
---

started: Alexey Larionov, 2016  
last updated: Alexey Larionov, 03Apr2017

# Summary

Eigenvectors are calculated using **42,250** common variants only:  
5% < AF  < 95% in each of the compared datasets (WECARE and NFE)  

Overall, eigenvectors are calculated for 3 datasets:  

- wecare-only
- wecare-nfe (using overlap wecare and nfe)
- wecare-nfe-50kgen (using overlap wecare, nfe and kgen50)  

This script deals with wecare-nfe dataset 

Requires f01_calculate_eigenvectors.R  

Input data: 275,516 vars x 728 cases (480 wecare + 198 nfe + 50 kgen50)  
Output data: 275,516 vars x 678 cases (480 wecare + 198 nfe)  

# start_section

```{r start_section}

# Time stamp
Sys.time()

# Folders
#/analysis/mtgroup_share
#/scratch/medgen
setwd("/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts")
source_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/source_data"
interim_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/interim_data"
results_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/results"

# Required libraries
library(ggplot2)
library(plotly)

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "r05_add_kgen50_Apr2017.RData", sep="/"))

```

# check_data

728 samples = 480 wecare + 198 nfe + 50 kgen50  

```{r check_data}

ls()

dim(genotypes.mx)
class(genotypes.mx)
genotypes.mx[1:5, 1:5]

dim(phenotypes.df)
str(phenotypes.df)
phenotypes.df[1:5, 1:5]

dim(variants.df)
str(variants.df)
variants.df[1:5,1:5]

dim(kgen.df)
str(kgen.df)
kgen.df[1:5,1:5]

dim(exac.df)
str(exac.df)
exac.df[1:5,1:5]

# Check consistence of rownames in kgen50_gt.mx and kgen50_vcf.df
sum(rownames(genotypes.mx) != rownames(variants.df))
sum(rownames(genotypes.mx) != rownames(kgen.df))
sum(rownames(genotypes.mx) != rownames(exac.df))

sum(colnames(genotypes.mx) != rownames(phenotypes.df))

```

# make_wecare_nfe_subset_of_data

keep 678 wecare-nfe cases: 480 wecare + 198 nfe

```{r make_wecare_only_subset_of_data}

# Remove nfe cases
wecare_nfe_cases <- as.vector(phenotypes.df[phenotypes.df$cc > -2,"wes_id"])
length(wecare_nfe_cases) # 678

wecare_nfe_genotypes.mx <- genotypes.mx[,wecare_nfe_cases]
dim(wecare_nfe_genotypes.mx) # 275,516 x 678
wecare_nfe_genotypes.mx[1:5,1:5]

wecare_nfe_phenotypes.df <- phenotypes.df[wecare_nfe_cases,]
dim(wecare_nfe_phenotypes.df) # 678 x 32
wecare_nfe_phenotypes.df[1:5,1:5]

wecare_nfe_variants.df <- variants.df
wecare_nfe_exac.df <- exac.df
wecare_nfe_kgen.df <- kgen.df

# Clean-up
rm(wecare_nfe_cases, genotypes.mx, phenotypes.df, variants.df, exac.df, kgen.df)

```

# check_for_variants_with_uniform_genotypes_accross_all_wecare_nfe_samples

There should be no such variants because they were removed earlier,  
when generated wecare-nfe dataset

```{r check_for_variants_with_uniform_genotypes_accross_all_wecare_nfe_samples}

# Check that there is no all-NA variants 
# (this is assumed for some downstream calculations)
non_NA_count.udf <- function(x){sum(!is.na(x))}
all_NA <- apply(wecare_nfe_genotypes.mx, 1, non_NA_count.udf) == 0
sum(all_NA) # 0

# Function to detect uniform numeric vector
uniform_vector.udf <- function(x){
  if(min(x, na.rm=TRUE) == max(x, na.rm=TRUE)){return(TRUE)} else {return(FALSE)}}

# Variants with uniform genotypes accross all samples 
uniform_genotypes <- apply(wecare_nfe_genotypes.mx, 1, uniform_vector.udf)
summary(uniform_genotypes)
sum(uniform_genotypes) # 0

# Clean-up
rm(non_NA_count.udf, all_NA, uniform_vector.udf, uniform_genotypes)

```

# recalculate_total_AFs

Used for selecting variants for eigenvectors computation.  
Could be recalculated later after exclusion of eigenvectors outliers.  

```{r recalculate_total_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Rename AF fields in the variants table
vars_colnames <- colnames(wecare_nfe_variants.df)
"ac_raw" -> vars_colnames[ vars_colnames == "AC" ]
"an_raw" -> vars_colnames[ vars_colnames == "AN" ]
"af_raw" -> vars_colnames[ vars_colnames == "AF" ]
vars_colnames -> colnames(wecare_nfe_variants.df)

# --- Calculate total AFs --- #

ac_all <- apply(wecare_nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_all <- apply(wecare_nfe_genotypes.mx, 1, get_allele_number.udf)
af_all <- ac_all/an_all

# Ceck AFs 
# (note that uniform variants were excluded)
ac_all[1:10]
an_all[1:10]
af_all[1:10]

min(ac_all)
min(an_all)
min(af_all)

max(ac_all)
max(an_all)
max(af_all)

# Add new AFs to the variants table
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_all, an_all, af_all)

# Clean-up
rm(vars_colnames, ac_all, an_all, af_all)

```

# calculate_nfe_AFs

```{r calculate_nfe_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
nfe_cases <- wecare_nfe_phenotypes.df$cc == -1 
sum(nfe_cases) # 198
nfe_genotypes.mx <- wecare_nfe_genotypes.mx[,nfe_cases]
dim(nfe_genotypes.mx)

# Calculate ubc ac, an and af
ac_nfe <- apply(nfe_genotypes.mx, 1, sum, na.rm=TRUE)
an_nfe <- apply(nfe_genotypes.mx, 1, get_allele_number.udf)
af_nfe <- ac_nfe/an_nfe

# Add updated AFs to wecare_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_nfe, an_nfe, af_nfe)

# Clean-up
rm(get_allele_number.udf, nfe_cases, nfe_genotypes.mx, ac_nfe, an_nfe, af_nfe)

```

# calculate_wecare_AFs

```{r calculate_wecare_AFs}

# Function to calculate AN
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# Prepare genotypes table
wecare_cases <- wecare_nfe_phenotypes.df$cc > -1 
sum(wecare_cases) # 480
wecare_genotypes.mx <- wecare_nfe_genotypes.mx[,wecare_cases]
dim(wecare_genotypes.mx)

# Calculate ubc ac, an and af
ac_wecare <- apply(wecare_genotypes.mx, 1, sum, na.rm=TRUE)
an_wecare <- apply(wecare_genotypes.mx, 1, get_allele_number.udf)
af_wecare <- ac_wecare/an_wecare

# Add updated AFs to wecare_variants.df
wecare_nfe_variants.df <- cbind(wecare_nfe_variants.df, ac_wecare, an_wecare, af_wecare)

# Clean-up
rm(get_allele_number.udf, wecare_cases, wecare_genotypes.mx, ac_wecare, an_wecare, af_wecare)

```

# calculate_eigenvectors

Only common variants will be used for eigenvectors calculation.  

Note exclusion on both sides: low- and high- AFs:  
- Low AFs remove rare variants with common allele in reference genome  
- Hight AFs remove rare variants with common allele in reference genome  

```{r calculate_eigenvectors}

# --- Make subset of data for eigenvectors calculation --- #

# Select wecare-nfe overlap variants

nfe_common_vars <- wecare_nfe_variants.df$af_nfe > 0.05 & wecare_nfe_variants.df$af_nfe < 0.95
sum(nfe_common_vars) # 45,232

wecare_common_vars <- wecare_nfe_variants.df$af_wecare > 0.05 & wecare_nfe_variants.df$af_wecare < 0.95
sum(wecare_common_vars) # 44,524

common_overlap_vars <- nfe_common_vars & wecare_common_vars
sum(common_overlap_vars) # 42,250

# Make the subset of data
common_overlap_wecare_nfe_genotypes.mx <- wecare_nfe_genotypes.mx[common_overlap_vars,]
dim(common_overlap_wecare_nfe_genotypes.mx)
common_overlap_wecare_nfe_genotypes.mx[1:5,1:5]

# --- Calculate eigenvectors --- #

source("f01_calculate_eigenvectors.R")
wecare_nfe.eigen <- normalise_and_calculate_eigenvectors.udf(common_overlap_wecare_nfe_genotypes.mx)

# Clean-up
rm(common_overlap_vars, wecare_common_vars, nfe_common_vars, cbc_common_vars, ubc_common_vars, 
   normalise_and_calculate_eigenvectors.udf, common_overlap_wecare_nfe_genotypes.mx)

```

# plot_eigenvectors

```{r plot_eigenvectors}

# --- Prepare data for plotting --- #

wecare_nfe.eigenvectors.df <- as.data.frame(wecare_nfe.eigen$vectors) # eigenvectors in columns

# Prepare cases lables
cases_labels <- as.vector(wecare_nfe_phenotypes.df$cc)
"NFE" -> cases_labels[cases_labels==-1]
"UBC" -> cases_labels[cases_labels==0]
"CBC" -> cases_labels[cases_labels==1]

summary(as.factor(cases_labels))

# Cases IDs (for labels on interactive plot)
cases_IDs <- as.vector(wecare_nfe_phenotypes.df$wes_id)

# Prepare colour scale
colours <- c("NFE" = "BLUE", "UBC" = "PINK", "CBC" = "RED")
userColourScale <- scale_colour_manual(values=colours)

# Data frame to plot
data2plot.df <- cbind(cases_IDs, cases_labels, wecare_nfe.eigenvectors.df[,1:5])

# --- Plot eig1 vs eig2 --- #

g <- ggplot(data2plot.df, aes(V1,V2)) +
  geom_point(aes(colour=cases_labels, text=cases_IDs)) + 
  labs(title="wecare-nfe<br>common overlap variants (42,250 x 678)", x="eigenvector1", y="eigenvector2") +
  userColourScale

ggplotly(g)

# --- Plot eig2 vs eig3 --- #

g <- ggplot(data2plot.df, aes(V2,V3)) +
  geom_point(aes(colour=cases_labels, text=cases_IDs)) + 
  labs(title="wecare-nfe<br>common overlap variants (42,250 x 678)", x="eigenvector2", y="eigenvector3") +
  userColourScale

ggplotly(g)

# --- Plot eig3 vs eig4 --- #

g <- ggplot(data2plot.df, aes(V3,V4)) +
  geom_point(aes(colour=cases_labels, text=cases_IDs)) + 
  labs(title="wecare-nfe<br>common overlap variants (42,250 x 678)", x="eigenvector3", y="eigenvector4") +
  userColourScale

ggplotly(g)

# --- Plot eig4 vs eig5 --- #

g <- ggplot(data2plot.df, aes(V4,V5)) +
  geom_point(aes(colour=cases_labels, text=cases_IDs)) + 
  labs(title="wecare-nfe<br>common overlap variants (42,250 x 678)", x="eigenvector4", y="eigenvector5") +
  userColourScale

ggplotly(g)

# --- Clean-up --- #

rm(wecare_nfe.eigenvectors.df, g, data2plot.df, cases_IDs, cases_labels, colours, userColourScale)

```

# calculate_eigenvector_outliers_for_wecare_nfe

Explore 6 standard deviations in 5 top eigenvectors  

```{r calculate_eigenvector_outliers_for_wecare_nfe}

wecare_nfe_eigenvectors.mx <- wecare_nfe.eigen$vectors

ev1 <- wecare_nfe_eigenvectors.mx[,1]
ev1.positive_outliers <- ev1 > mean(ev1) + 6 * sd(ev1)
ev1.negative_outliers <- ev1 < mean(ev1) - 6 * sd(ev1)
sum(ev1.positive_outliers)
sum(ev1.negative_outliers)

wecare_nfe_phenotypes.df$wes_id[ev1.positive_outliers]
wecare_nfe_phenotypes.df$wes_id[ev1.negative_outliers]

ev2 <- wecare_nfe_eigenvectors.mx[,2]
ev2.positive_outliers <- ev2 > mean(ev2) + 6 * sd(ev2)
ev2.negative_outliers <- ev2 < mean(ev2) - 6 * sd(ev2)
sum(ev2.positive_outliers)
sum(ev2.negative_outliers)

wecare_nfe_phenotypes.df$wes_id[ev2.positive_outliers]
wecare_nfe_phenotypes.df$wes_id[ev2.negative_outliers]

ev3 <- wecare_nfe_eigenvectors.mx[,3]
ev3.positive_outliers <- ev3 > mean(ev3) + 6 * sd(ev3)
ev3.negative_outliers <- ev3 < mean(ev3) - 6 * sd(ev3)
sum(ev3.positive_outliers)
sum(ev3.negative_outliers)

wecare_nfe_phenotypes.df$wes_id[ev3.positive_outliers]
wecare_nfe_phenotypes.df$wes_id[ev3.negative_outliers]

ev4 <- wecare_nfe_eigenvectors.mx[,4]
ev4.positive_outliers <- ev4 > mean(ev4) + 6 * sd(ev4)
ev4.negative_outliers <- ev4 < mean(ev4) - 6 * sd(ev4)
sum(ev4.positive_outliers)
sum(ev4.negative_outliers)

wecare_nfe_phenotypes.df$wes_id[ev4.positive_outliers]
wecare_nfe_phenotypes.df$wes_id[ev4.negative_outliers]

ev5 <- wecare_nfe_eigenvectors.mx[,5]
ev5.positive_outliers <- ev5 > mean(ev5) + 6 * sd(ev5)
ev5.negative_outliers <- ev5 < mean(ev5) - 6 * sd(ev5)
sum(ev5.positive_outliers)
sum(ev5.negative_outliers)

wecare_nfe_phenotypes.df$wes_id[ev5.positive_outliers]
wecare_nfe_phenotypes.df$wes_id[ev5.negative_outliers]

# Clean-up
rm(wecare_nfe_eigenvectors.mx, 
   ev1, ev1.positive_outliers, ev1.negative_outliers, 
   ev2, ev2.positive_outliers, ev2.negative_outliers, 
   ev3, ev3.positive_outliers, ev3.negative_outliers,
   ev4, ev4.positive_outliers, ev4.negative_outliers,
   ev5, ev5.positive_outliers, ev5.negative_outliers)

```

# data_summary

```{r data_summary}

ls()

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(wecare_nfe_variants.df)
colnames(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

str(wecare_nfe.eigen)

sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "r06b_calculate_egenvectors_wecare_nfe_Apr2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
