---
title: "QQ-plots nfe vs wecare"
output: html_document
---

started: Alexey Larionov, 25Mar2017  
last updated: Alexey Larionov, 28Mar2017  

# Summary

Make QQ plot for simple logit nfe vs wecare to see whether there is an inflation 
(recommended by DC during the Skype meeting of 14Mar2017)

# start_section

```{r start_section}

# Start time
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_03.17/scripts")
source_data_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/source_data"
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/interim_data"
results_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/results"

# Load function for nice qq-plots
source("f03_qqplot.R")

```

# load_data

```{r load_data}

# Load data
load(paste(interim_data_folder, "s07b_filter_by_variant_effect_wecare_nfe_Mar2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(wecare_nfe_variants.df)
colnames(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

# Check consistency of rownames and colnames
sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))

```

# prepare_data_for_analysis

No missed data in eigenvectors or outcomes  
Imputed missed data for genotypes  

```{r prepare_data_for_analysis}

# Eigenvectors
E <- wecare_nfe.eigen$vectors[,1:3]
sum(is.na(E)) # 0

# Outcomes
Y <- ifelse(wecare_nfe_phenotypes.df$cc==-1, 0, 1)
sum(is.na(Y)) # 0

# Genotypes: select variants with AF >5%, present in both wecare and nfe
weccare_common_vars <- wecare_nfe_variants.df$AF_wecare > 0.05
nfe_common_vars <- wecare_nfe_variants.df$AF_nfe > 0.05
wecare_nfe_common_vars <- weccare_common_vars & nfe_common_vars
sum(wecare_nfe_common_vars) # 1,079

# Genotypes: transpose and impute missed data
G <- t(wecare_nfe_genotypes.mx[wecare_nfe_common_vars,])
G <- apply(G, 2, function(v) { ifelse(is.na(v), mean(v, na.rm=T), v) })
sum(is.na(G)) # 0

```

# logit_regression

Reccurent warnings:  
Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred  
Warning: glm.fit: algorithm did not converge  

```{r logit_regression, warning=FALSE}

# Write header to output matrix 
# (do not use data frame here: it will play with rbind!)
result.mx <- matrix(ncol=3, nrow=0) 
colnames(result.mx) <- c("VarID", "pX", "pXE")

wecare_nfe_common_variants.df <- wecare_nfe_variants.df[wecare_nfe_common_vars,]

# For each variant
for(v in 1:nrow(wecare_nfe_common_variants.df)){
  
  # Get variant ID
  var_id <- as.vector(wecare_nfe_common_variants.df$SplitVarID[v])

  # Get vector of genotypes  
  X <- G[,v]

  # Calculate logit regression
  regX <- glm(Y ~ X, family=binomial)
  regXE <- glm(Y ~ X + E, family=binomial)

  # Get p-values
  pX <- summary(regX)$coef["X", "Pr(>|z|)"]
  pXE <- summary(regXE)$coef["X", "Pr(>|z|)"]
  
  # Calculate p-value estimate using LRT?
  #regE <- glm(Y ~ E, family=binomial)
  #chi_stat_lrt = 2*(logLik(regXE) - logLik(regE))
  #p_lrt = 1-pchisq(chi_stat_lrt, df=1) # df??

  # how to get an alteranative p-value from anova(reg) ?
  
  # Record output to results table (p being converted to char?)
  result.mx <- rbind(result.mx, c(var_id, pX, pXE))
  
} # next variant

# Clean-up
rm(X, Y, G, E, v, var_id, regX, regXE, pX, pXE)

```

explore_result

```{r explore_result}

library(qqman)
result.df <- as.data.frame(result.mx)
result.df$pX <- as.numeric(as.vector(result.df$pX))
result.df$pXE <- as.numeric(as.vector(result.df$pXE))

hist(result.df$pX)
hist(result.df$pXE)

qq(result.df$pX)
qq(result.df$pXE)

qqunif.plot(result.df$pX, main="WECARE vs NFE: w/o eigenvectors")

```
