---
title: "variants_logit_wecare_only_Mar2017"
output: html_document
---

started: Alexey Larionov, 26Mar2017  
last updated: Alexey Larionov, 27Mar2017  

# Summary

Make QQ plot for variants simple logit cbc vs ubc:  
to see whether there is an inflation and how the methods work  

# start_section

```{r start_section}

# Start time
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_03.17/scripts")
source_data_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/source_data"
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/interim_data"
results_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/results"

```

# load_data

```{r load_data}

# Load data
load(paste(interim_data_folder, "s07a_filter_by_variant_effect_wecare_only_Mar2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_genotypes.mx)
class(wecare_genotypes.mx)
wecare_genotypes.mx[1:5,1:5]

dim(wecare_phenotypes.df)
str(wecare_phenotypes.df)
wecare_phenotypes.df[1:5,1:5]

dim(wecare_variants.df)
colnames(wecare_variants.df)
wecare_variants.df[1:5,1:5]

dim(wecare_kgen.df)
colnames(wecare_kgen.df)
wecare_kgen.df[1:5,1:5]

dim(wecare_exac.df)
colnames(wecare_exac.df)
wecare_exac.df[1:5,1:5]

# Check consistency of rownames and colnames
sum(colnames(wecare_genotypes.mx) != rownames(wecare_phenotypes.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_variants.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_kgen.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_exac.df))

```

# prepare_data_for_analysis

No missed data in outcome or eigenvectors  
Impute missed data for genotypes  

```{r prepare_data_for_analysis}

# Outcomes
Y <- wecare_phenotypes.df$cc
sum(is.na(Y)) # 0
#length(Y)
#sum(Y)

# Genotypes: transpose non-aggregated variants and impute missed data
G <- t(wecare_genotypes.mx)
G <- apply(G, 2, function(v) { ifelse(is.na(v), mean(v, na.rm=T), v) })
sum(is.na(G)) # 0

# Eigenvectors
E <- wecare.eigen$vectors[,c(1:3)]  
sum(is.na(E)) # 0

```

# logit_regression

< 10 min  
warnings=FALSE  
Recurrent warning: glm.fit: fitted probabilities numerically 0 or 1 occurred  

```{r logit_regression, warning=FALSE}

# Write header to output matrix 
# (do not use data frame here: it will play with rbind!)
result.mx <- matrix(ncol=4, nrow=0) 
colnames(result.mx) <- c("var_id", "p_x", "p_xe", "p_lrt")

# For each variant
for(v in 1:nrow(wecare_variants.df)){
  
  # Get variant ID
  var_id <- as.vector(wecare_variants.df$SplitVarID[v])

  # Get vector of genotypes  
  X <- G[,v]

  # Calculate logit regression
  reg_x <- glm(Y ~ X, family=binomial)
  reg_e <- glm(Y ~ E, family=binomial)
  reg_xe <- glm(Y ~ X + E, family=binomial)

  # Get p-values
  p_x <- summary(reg_x)$coef["X", "Pr(>|z|)"]
  p_xe <- summary(reg_xe)$coef["X", "Pr(>|z|)"]
  
  # Calculate p-value estimate using LRT?
  chi_stat_lrt = 2*(logLik(reg_xe) - logLik(reg_e))
  p_lrt = 1 - pchisq(chi_stat_lrt, df=1) # df??

  # Additionally: get an alteranative p-value from anova(reg) ??
  # anova(reg_x, reg_xe) ...
  
  # Record output to results table (p being converted to char)
  result.mx <- rbind(result.mx, c(var_id, p_x, p_xe, p_lrt))
  
} # next variant

# Clean-up
rm(X, Y, G, E, v, var_id, reg_x, reg_e, reg_xe, chi_stat_lrt, p_x, p_xe, p_lrt)

```

# explore_result

```{r explore_result}

# Check 
dim(result.mx)

# Cojnvert p estimates to numeric
result.df <- as.data.frame(result.mx)
result.df$p_x <- as.double(as.vector(result.df$p_x))
result.df$p_xe <- as.double(as.vector(result.df$p_xe))
result.df$p_lrt <- as.double(as.vector(result.df$p_lrt))

# Sort
result.df <- result.df[order(result.df$p_lrt),]

# --- Compare p-estimates --- #

plot(result.df$p_x, result.df$p_xe, 
     main="Simple glm, +/ eigenvectors", 
     xlab="No eigenvectors", ylab="With 3 eigenvectors")
abline(0,1,col="red",lwd=3)

plot(result.df$p_xe, result.df$p_lrt, 
     main="Simple glm vs lrt, (with 3 eigenvectors)",
     xlab="glm", ylab="lrt")
abline(0,1,col="red",lwd=3)

# --- QQ-plots --- #

source("f03_qqplot.R")
qqunif.plot(result.df$p_xe, main="CBC vs UBC, 3 eigenvectors (simple glm)")
qqunif.plot(result.df$p_lrt, main="CBC vs UBC, 3 eigenvectors (LRT)")

# Clean-up
rm(result.mx, qqunif.plot)

```

# save_results

```{r save_results}

save.image(paste(results_folder, "r08a_variants_logit_wecare_only_Mar2017.RData", sep="/"))

write.table(result.df,
            file=paste(results_folder, "r08a_variants_logit_wecare_only_Mar2017.txt", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
