---
title: "matched_analysis_wecare_only_Mar2017"
output: html_document
---

started: Alexey Larionov, 27Feb2017  
last updated: Alexey Larionov, 28Mar2017

# Summary

Reads and checks data.  

Checks for control's AFs > 0.5, inverts genotypes and recalculates AFs if needed.  

Weights variants by AFs in controls (ubc) and aggregats variants by gene.  

Then performs matched analysis and compares it with un-matched glm (based on David Conti e-mail of 18Nov2016).    

Calculates aggregated AFs in all sub-groups (ubc and cbc +/- fam), for convinience.  

Writes results into a text file.  

# start_section

```{r start_section}

# Start time
Sys.time()

# Folders
setwd("/scratch/medgen/scripts/wecare_stat_03.17/scripts")
source_data_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/source_data"
interim_data_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/interim_data"
results_folder <- "/scratch/medgen/scripts/wecare_stat_03.17/results"

# Load libraries and functions
library(survival) # for clogit (matched analysis)
library(VennDiagram)
library(ggplot2)

source("f03_qqplot.R") # for nice qq-plots

```

# load_data

```{r load_data}

# Load data
load(paste(interim_data_folder, "s07a_filter_by_variant_effect_wecare_only_Mar2017.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(wecare_genotypes.mx)
class(wecare_genotypes.mx)
wecare_genotypes.mx[1:5,1:5]

dim(wecare_phenotypes.df)
str(wecare_phenotypes.df)
wecare_phenotypes.df[1:5,1:5]

dim(wecare_variants.df)
colnames(wecare_variants.df)
wecare_variants.df[1:5,1:5]

dim(wecare_kgen.df)
colnames(wecare_kgen.df)
wecare_kgen.df[1:5,1:5]

dim(wecare_exac.df)
colnames(wecare_exac.df)
wecare_exac.df[1:5,1:5]

# Check consistency of rownames and colnames
sum(colnames(wecare_genotypes.mx) != rownames(wecare_phenotypes.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_variants.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_kgen.df))
sum(rownames(wecare_genotypes.mx) != rownames(wecare_exac.df))

```

# prepare_data_for_analysis

No missed data in any covariate, outcome or supplementary info 

```{r prepare_data_for_analysis}

# Eigenvectors
E <- wecare.eigen$vectors[,1:3]
sum(is.na(E)) # 0

# Matching variables
M <- as.matrix(wecare_phenotypes.df[,c("age_dx", "rstime")])  
sum(is.na(M)) # 0

# Matched pairs IDs
pairID <- wecare_phenotypes.df$setno
sum(is.na(pairID))

# Make a variable for hormone treatment.
hormone <- wecare_phenotypes.df$hormone

# Make variable for cytotoxic treatment
chemo <- wecare_phenotypes.df$chemo_cat != "no"

# Make joined treatment variable
treatment <- as.integer(chemo | hormone)
sum(is.na(treatment))

# Number of pregnancies
num_preg <- as.integer(wecare_phenotypes.df$num_preg)
sum(is.na(num_preg))

# Bind clinical covariates together
C <- cbind(treatment, num_preg)

# Outcomes
Y <- wecare_phenotypes.df$cc
sum(is.na(Y))

# Genes
genes <- unique(as.vector(wecare_variants.df$SYMBOL))
sum(is.na(genes))

# Clean-up
rm(chemo, hormone, treatment, num_preg)

```

# calculate_ubc_cbc_AFs

```{r calculate_ubc_cbc_AFs}

# Function to count alleles (2x for diploid)
get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# --- calculate_ubc_AFs --- #

# Prepare genotypes table
ubc_cases <- wecare_phenotypes.df$cc == 0 
ubc_genotypes.mx <- wecare_genotypes.mx[,ubc_cases]

# Calculate ubc ac, an and af
ac_ubc_cln <- apply(ubc_genotypes.mx, 1, sum, na.rm=TRUE)
an_ubc_cln <- apply(ubc_genotypes.mx, 1, get_allele_number.udf)
af_ubc_cln <- ac_ubc_cln/an_ubc_cln

# Add updated AFs to wecare_variants.df
wecare_variants.df <- cbind(wecare_variants.df, 
  ac_ubc_cln, an_ubc_cln, af_ubc_cln)

# Clean-up
rm(ubc_cases, ubc_genotypes.mx, ac_ubc_cln, an_ubc_cln, af_ubc_cln)

# --- calculate_cbc_AFs --- #

# Prepare genotypes table
cbc_cases <- wecare_phenotypes.df$cc == 1 
cbc_genotypes.mx <- wecare_genotypes.mx[,cbc_cases]

# Calculate cbc ac, an and af
ac_cbc_cln <- apply(cbc_genotypes.mx, 1, sum, na.rm=TRUE)
an_cbc_cln <- apply(cbc_genotypes.mx, 1, get_allele_number.udf)
af_cbc_cln <- ac_cbc_cln/an_cbc_cln

# Add updated AFs to wecare_variants.df
wecare_variants.df <- cbind(wecare_variants.df, 
  ac_cbc_cln, an_cbc_cln, af_cbc_cln)

# Clean-up
rm(cbc_cases, cbc_genotypes.mx, ac_cbc_cln, an_cbc_cln, af_cbc_cln)

```

# prepare_output_table

```{r prepare_output_table}

# Raw AFs headers
raw_afs_cols <- c(
    "ac_ubc_cln", "an_ubc_cln", "af_ubc_cln", 
    "ac_cbc_cln", "an_cbc_cln", "af_cbc_cln")

# Output AFs headers (w/o "_cln")
afs_cols <- sub("_cln","",raw_afs_cols)

# write header to output file
header <- c("gene", "n_variants", "multilaalelic_varaints", "inverted_afs", 
            "clogit_mode", "p_clogit", "p_clogit_anova", 
            "glm_mode", "p_glm", "p_glm_lrt", afs_cols)

write(paste(header, sep="", collapse="\t"),
  paste(results_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.txt", sep="/"))

# Clean-up
rm(header)

```

# do_analysis

Prepare output table, then
loop through each gene: 
- get varaints per gene  
- get tables with varaints and genotypes data  
- check and correct variants with inverted AFs (varaints with ubc ALT AF > 0.5)  
- recalculate AFs (will differ in the inverted varaints)  
- calculate weights and weighted predictive varaible
- perform matched and non-matched analyses  
- write results to text table  

Some errors and warnings are generated by clogit / glm functions.  
They are handled with try() and reported in clogit_mode / glm_mode columns

Chunk options: error=FALSE, warning=FALSE  
to suppress clattering with errors/warnings

```{r do_analysis, error=FALSE, warning=FALSE}

# Function to count alleles (2x for diploid)
#get_allele_number.udf <- function(x){2*sum(!is.na(x))}

# --- Loop through each gene --- #

# Genes for testing:
#gene="ATM" # many variants per gene
#gene="CASP9" # one variant per gene 
#gene="ADH1B" # contains a rare version of SNP in ref genome (2 vars altogether)
#gene="SLC27A3" # contains a multiallelic SNP (5 vars altogether)
#gene="CLC9A3R1" # a candidate from previous analyses
#gene="CCDC27" # failed fitting clogit

#genes <- c("ATM", "CASP9", "ADH1B", "SLC27A3", "SLC9A3R1", "CCDC27")

for(gene in genes){
  
  #gene="ATM" # 15 variants
  #gene="GABRD" # 1 variant
  
  # --- Select varaints data --- #
  
  # Get variants
  vars <- as.vector(wecare_variants.df[wecare_variants.df$SYMBOL==gene, "SplitVarID"])
  num_vars <- length(vars)
  var_IDs <- as.vector(wecare_variants.df[vars,"SplitVarID"])
    
  # Notes: 
  
  # I explicitly specify the class (i.e. data frame or matrix) and the num of rows.
  # This is done to ensure that single-variant genes are proecessed as expected; 
  # otherwise they might be converted to vectors etc. 
  
  # It seems that rownames and colnames are preserved, 
  # so I did no explicit provision to assure this. 
  variants.df <- as.data.frame(wecare_variants.df[vars,], nrow=num_vars)

  # Flag genes containing multiallelics varaints
  multiallelic_varaints <- FALSE
  if(sum(variants.df$"Multiallelic") > 0){
      multiallelic_varaints <- TRUE
  }

  # Raw allelic frequencies and counts
  raw_afs.df <- as.data.frame(variants.df[, raw_afs_cols])
  colnames(raw_afs.df) <- afs_cols # remove tne "_cln" suffic from AFs colnames

  # Raw AFs in controls (ubc)
  raw_afs_in_controls <- variants.df$af_ubc_cln

  # --- Raw genotypes --- #
  
  # Notes: 
  
  # I explicitly specify the class (i.e. data frame or matrix) and the num of rows. 
  # This is done to ensure that singel-variant genes are proecessed as expected; 
  # otherwise they might be converted to vectors etc. 
  
  # It seems that rownames and colnames may NOT be preserved at this occasion, 
  # so I explicitly assigned them too.  

  raw_genotypes.mx <- matrix(wecare_genotypes.mx[vars,], nrow=num_vars)
  rownames(raw_genotypes.mx) <- var_IDs
  colnames(raw_genotypes.mx) <- colnames(wecare_genotypes.mx)

  # Assume no inverted AFs in controls (no controls AFs > 0.5)
  inverted_afs <- FALSE
  afs_in_controls <- raw_afs_in_controls
  genotypes.mx <- raw_genotypes.mx
  afs.df <- raw_afs.df
  
  ##############################################################
  #             Check and correct for inverted AFs             #
  ##############################################################
  
  # For each var in the gene
  for(var in 1:num_vars){
    
    # Check whether this variant has inverted controls AF
    if(raw_afs_in_controls[var] > 0.5){
      
      #--------------------------------------------------------#
      #                Set flag for inverted_afs               #
      #--------------------------------------------------------#
      
      inverted_afs <- TRUE
      
      #--------------------------------------------------------#
      #      Update genotypes (swop 0 and 2 in genotypes)      #
      #--------------------------------------------------------#
      
      raw_var_genotypes <- raw_genotypes.mx[var,]
      var_genotypes <- raw_genotypes.mx[var,]
      0 -> var_genotypes[raw_var_genotypes == 2]
      2 -> var_genotypes[raw_var_genotypes == 0]
      var_genotypes -> genotypes.mx[var,]
      
      # Clean-up
      rm(raw_var_genotypes)
      
      #--------------------------------------------------------#
      #                     Recalculate AFs                    #
      #--------------------------------------------------------#
      
      ubc_cases <- wecare_phenotypes.df$cc == 0 
      afs.df[var,"ac_ubc"] <- sum(var_genotypes[ubc_cases], na.rm=TRUE)
      afs.df[var,"an_ubc"] <- get_allele_number.udf(var_genotypes[ubc_cases])
      afs.df[var,"af_ubc"] <- afs.df[var,"ac_ubc"]/afs.df[var,"an_ubc"]

      cbc_cases <- wecare_phenotypes.df$cc == 1 
      afs.df[var,"ac_cbc"] <- sum(var_genotypes[cbc_cases], na.rm=TRUE)
      afs.df[var,"an_cbc"] <- get_allele_number.udf(var_genotypes[cbc_cases])
      afs.df[var,"af_cbc"] <- afs.df[var,"ac_cbc"]/afs.df[var,"an_cbc"]

      #--------------------------------------------------------#
      #            Check and update the control's af           #
      #--------------------------------------------------------#
      
      if( afs.df[var, "af_ubc"] + raw_afs_in_controls[var] != 1 ){
        stop(paste("Error during AFs inversion, gene:", gene))
      } # there could be rounding diferences in order of 10E-15
      
      afs_in_controls[var] <- afs.df[var, "af_ubc"] 
      
      # Clean-up
      rm(ubc_cases, cbc_cases, var_genotypes)

    } # --- done the genotypes inversion for a variant (if needed) --- #

  } # -------------- Check next variant for the gene ----------------- #

  # Clean-up
  rm(raw_afs_in_controls, raw_genotypes.mx, raw_afs.df, var)

  ####################################################################################
  #                   Get aggregated allelic counts and fractions                    #
  ####################################################################################

  afs <- apply(afs.df, 2, sum, na.rm=TRUE)
  afs["af_ubc"] <- afs["ac_ubc"] / afs["an_ubc"]
  afs["af_cbc"] <- afs["ac_cbc"] / afs["an_cbc"]

  # Clean-up
  rm(afs.df)
  
  ####################################################################################
  #                     Prepare predictive variable for analyses                     #
  ####################################################################################

  # Transpose genotypes (make variants in columns)
  g <- t(genotypes.mx)
  
  # Impute missed genotypes (variant-wise means)
  g <- apply(g, 2, function(v) { ifelse(is.na(v), mean(v, na.rm=T), v) })

  # Calculate weights (the same as SKAT's default weights)
  w <- dbeta(afs_in_controls, 1, 25)

  # Calculate the predictive variable
  # sum of weighted ACs per case
  # as n x 1 matrix
  X <- g%*%w    
  
  # Make the variable numewric vector 
  # (not a matrix etc - just in case:
  # a matrix still would work ...)
  X_names  <- rownames(X)
  X <- as.numeric(X)
  names(X) <- X_names
  
  # Clean-up
  rm(genotypes.mx, g, w, X_names)

  ####################################################################################
  # Burden type analysis using conditional logistic regression with a LRT and weight #
  ####################################################################################
  
  # requires library(survival)
  # What is about the presense of unmatched? 
  # What is about absense of covatiates in DC's example? 
  
  # exploring reg and anova(reg) objects:
  # 2-nd eigenvector is close to significance ...
  # Treatment is close to significance ...
  # NumPreg is significant ...
  
  # Occasionally clogit fails with the following error+warning:
  
  # Error in fitter(X, Y, strats, offset, init, control, weights = weights,  : 
  # NA/NaN/Inf in foreign function call (arg 5)
  # In addition: Warning message:
  # In fitter(X, Y, strats, offset, init, control, weights = weights,  :
  # Ran out of iterations and did not converge

  # This has been discussed, for instance, here:
  # http://r.789695.n4.nabble.com/Error-when-running-Conditional-Logit-Model-td948556.html 
  
  # The below link advised using method="approximate" to work arpound the issue:
  # http://stats.stackexchange.com/questions/109222/conditional-logistic-regression-model-does-not-converge-but-logistic-regression
  
  # Try standard clogit
  clogit_mode <- "standard"
  try(clogit.reg <- 
        clogit( Y ~ X + E + C +
          strata(pairID)))
  
  # Try clogit with approximate calculations 
  # (if the standard clogit has failed)
  if( !exists("clogit.reg") ){
    
     clogit_mode <- "approximate"
     try(clogit.reg <- 
          clogit(Y ~ X + E + C +
            strata(pairID),
            method="approximate"))
  }

  # Calculate p-values from clogit.reg object
  if(exists("clogit.reg")){

    p.clogit <- summary(clogit.reg)$coef["X", "Pr(>|z|)"]
    
    try(p.clogit.anova <- anova(clogit.reg)["X","Pr(>|Chi|)"])

  }else{
    
    # If clogit has failed even with approximate method
    clogit_mode <- "failed_clogit"
    p.clogit <- NA
    p.clogit.anova <- NA

  }
  
  # If failed clogit anova
  if(! exists("p.clogit.anova")){

    # If clogit has failed even with approximate method
    clogit_mode <- "failed_anova"
    p.clogit.anova <- NA

  }

  
  # Clean-up
  rm(clogit.reg)

  ####################################################################################
  #                              Non-matched analysis                                #
  ####################################################################################

  # matching_variables <- as.matrix(wecare_phenotypes.df[,c("age_dx", "rstime")])  
  
  # Occasional warnings from glm (usually coinside with errors in clogit - see above)
  # glm.fit: fitted probabilities numerically 0 or 1 occurred 
  
  # Exploring reg objects:
  # 2nd eigenvector, treatment and num preg are significant 
  
  glm_mode <- "standard"
  
  # Data for glm models
  dat <- data.frame(Y, X, M)
  
  # Try null-model
  try(glm.reg.null <- 
        glm(Y ~ E + C +
            age_dx + rstime, 
            family=binomial, 
            data=dat))
  
  # Try complete-model
  try(glm.reg <- 
        glm(Y ~ X + E + C +
            age_dx + rstime, 
            family=binomial, 
            data=dat))

  # Calculate p-values from the glm models
  if( exists("glm.reg.null") & exists("glm.reg") ){

    # Read p-value from the glm model
    p.glm <- summary(glm.reg)$coef["X", "Pr(>|z|)"]
    
    # Calculate p-value estimate using LRT
    chi.stat.LRT = 2*(logLik(glm.reg) - logLik(glm.reg.null))
    p.glm.lrt = 1-pchisq(chi.stat.LRT, df=1)

  }else{
    
    # If clogit has failed even with approximate method
    glm_mode <- "failed"
    p.glm <- NA
    p.glm.lrt <- NA

  }

  # Clean-up
  rm(dat, glm.reg, glm.reg.null, chi.stat.LRT)
  
  ####################################################################################
  #                                  Write result                                    #
  ####################################################################################
  
  # Compile result
  # Header: "gene", "n_variants", "multilaalelic_varaints", "inverted_afs", 
  #         "clogit_mode", "p_clogit", "p_clogit_anova", 
  #         "glm_mode", "p_glm", "p_glm_lrt", 
  #          afs_cols
  result <- c(gene, num_vars, multiallelic_varaints, inverted_afs, 
              clogit_mode, p.clogit, p.clogit.anova, 
              glm_mode, p.glm, p.glm.lrt, 
              afs)
    
  # Record result to the output table
  write(paste(result, sep="", collapse="\t"),
    paste(results_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.txt", sep="/"),
    append = TRUE)

  # Clean-up
  rm(result, vars, variants.df, var_IDs, num_vars, multiallelic_varaints, inverted_afs, 
     clogit_mode, p.clogit, p.clogit.anova, glm_mode, p.glm, p.glm.lrt, afs, afs_in_controls, X)
  
} # next gene

# Clean-up
rm(gene, genes, Y, E, M, C, pairID, afs_cols, raw_afs_cols, get_allele_number.udf)

```

# read_and_sort_results_table

```{r read_and_sort_results_table}

# Read results
results.df <- read.table(paste(results_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.txt", sep="/"), header = TRUE)
dim(results.df)
str(results.df)

summary(as.factor(results.df$multilaalelic_varaints))
# FALSE  TRUE 
#  8463   303 

summary(as.factor(results.df$inverted_afs))
# FALSE  TRUE 
#  8618   148 

summary(as.factor(results.df$clogit_mode))
#  approximate  failed_anova failed_clogit      standard 
#          601            19            45          8,101 

summary(as.factor(results.df$glm_mode)) 
# standard: 8,766

# ------------ Compare P values ------------ #
mean_p_values <- c(
  mean(results.df$p_clogit, na.rm=TRUE),
  mean(results.df$p_clogit_anova, na.rm=TRUE),
  mean(results.df$p_glm),
  mean(results.df$p_glm_lrt))
names(mean_p_values) <- c("clogit", "clogit_anova", "glm", "glm_lrt")
barplot(mean_p_values)

colours <- as.vector(results.df$clogit_mode)
"black" -> colours[colours == "standard"]
"red" -> colours[colours == "approximate"]
"green" -> colours[colours == "failed_clogit"]
"green" -> colours[colours == "failed_anova"]

plot(p_glm_lrt ~ p_clogit_anova, data=results.df,
     col=colours, main="glm_lrt vs clogit_anova")
abline(a=0, b=1, col="blue", lwd=3)
legend("bottomright", inset=.05, 
       title="clogit mode:",
       c("standard", "approximate"), 
       fill=c("black", "red"),
       horiz = TRUE)

plot(p_clogit ~ p_clogit_anova, data=results.df, 
     col=colours, main="p_clogit vs clogit_anova")
abline(a=0, b=1, col="blue", lwd=3)
legend("bottomright", inset=.05, 
       title="clogit mode:",
       c("standard", "approximate"), 
       fill=c("black", "red"),
       horiz = TRUE)

plot(p_glm ~ p_glm_lrt, data=results.df, 
     col=colours, main="p_glm vs p_glm_lrt")
abline(a=0, b=1, col="blue", lwd=3)
legend("bottomright", inset=.05, 
       title="clogit mode:",
       c("standard", "approximate", "failed"), 
       fill=c("black", "red", "green"),
       horiz = TRUE)

# ---------------- qq plots ---------------- #

# http://genome.sph.umich.edu/wiki/Code_Sample:_Generating_QQ_Plots_in_R
# http://www.gettinggeneticsdone.com/2014/05/qqman-r-package-for-qq-and-manhattan-plots-for-gwas-results.html 

# It is belived that under null hypothesis the distribution of P-values shoulod be uniform
# e.g. http://physics.rockefeller.edu/luc/technical_reports/cdf8023_facts_about_p_values.pdf
# Hopefully, this holds for the GWAs-WES type of data

obs.pvalues <- as.numeric(results.df$p_glm_lrt)
exp.pvalues<-(rank(obs.pvalues, ties.method="first") + 0.5)/(length(obs.pvalues)+1)# # note the flat expectation
plot(-log10(exp.pvalues), -log10(obs.pvalues), 
     xlim = c(0,4.5), ylim = c(0,4.5),
     main="CBC vs UBC: p_glm_lrt")
abline(0,1,col="red")

qqunif.plot(obs.pvalues, main="CBC vs UBC: p_glm_lrt")

obs.pvalues <- as.numeric(results.df$p_clogit_anova[!is.na(results.df$p_clogit_anova)])
exp.pvalues<-(rank(obs.pvalues, ties.method="first") + 0.5)/(length(obs.pvalues)+1)
plot(-log10(exp.pvalues), -log10(obs.pvalues), 
     xlim = c(0,4.5), ylim = c(0,4.5),
     main="CBC vs UBC: p_clogit_anova")
abline(0,1,col="red")

qqunif.plot(obs.pvalues, main="CBC vs UBC: p_clogit_anova")

# --- Sort --- #

results.df <- results.df[order(as.numeric((results.df$p_glm_lrt))), ] 

# Save sorted
write.table(results.df, 
            file=paste(results_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.txt", sep="/"),
            row.names = FALSE, quote=FALSE, sep="\t")

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
