---
title: "nfe vs wecare: glm, no aggregating, no weighting, 3 ev nfe, QQ-plots"
output: html_document
---

started: Alexey Larionov, 25Mar2017  
last updated: Alexey Larionov, 03Apr2017  

# Summary

Make QQ plot(s) for nfe vs wecare to see whether there is an inflation 
(recommended by DC during the Skype meeting of 14Mar2017)

# start_section

```{r start_section}

# Start time
Sys.time()

# Folders
#/analysis/mtgroup_share
#/scratch/medgen
setwd("/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts")
source_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/source_data"
interim_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/interim_data"
results_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/results"

```

# load_data

```{r load_data}

# Load data
load(paste(interim_data_folder, "r07_pick_egenvectors_exclude_outliers_Apr2017.RData", sep="/"))

```

# check_data

```{r check_data}

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

dim(wecare_nfe_variants.df)
str(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

# Check consistency of rownames and colnames
sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))

sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))

```

# invert_genotypes

Invert genotypes for variants with total AF > 0.5  

```{r invert_genotypes}

# Prepare tables for output
wecare_nfe_variants_inv.df <- cbind(wecare_nfe_variants.df, 
                     inverted=rep(FALSE,nrow(wecare_nfe_variants.df)),
                     ac_inv=wecare_nfe_variants.df$ac_all,
                     an_inv=wecare_nfe_variants.df$an_all,
                     af_inv=wecare_nfe_variants.df$af_all)

wecare_nfe_genotypes_inv.mx <- wecare_nfe_genotypes.mx

# For each variant
for(var in 1:nrow(wecare_nfe_variants_inv.df)){
  
  # If total AF > 0.5
  if(wecare_nfe_variants_inv.df[var,"af_all"] > 0.5){
    
    # Flag variant
    TRUE -> wecare_nfe_variants_inv.df[var, "inverted"]
    
    # Invert genotypes 
    genotypes <- wecare_nfe_genotypes_inv.mx[var,]
    genotypes_inv <- genotypes
    0 -> genotypes_inv[genotypes == 2]
    2 -> genotypes_inv[genotypes == 0]
    genotypes_inv -> wecare_nfe_genotypes_inv.mx[var,]

    # Calculate inverted AFs
    ac_inv <- sum(genotypes_inv, na.rm=TRUE)
    an_inv <- 2*sum(!is.na(genotypes_inv))
    af_inv <- ac_inv/an_inv

    # Add inverted AFs to the variants data.frame
    wecare_nfe_variants_inv.df[var, c("ac_inv", "an_inv", "af_inv")] <- c(ac_inv, an_inv, af_inv)
    
    # Clean-up
    rm(genotypes, genotypes_inv, ac_inv, an_inv, af_inv)
    
  }
}

# Number of invered variants
sum(wecare_nfe_variants_inv.df$inverted) # 12,729

# Some effects of inversion
max(wecare_nfe_variants_inv.df$af_all) # ~0.99
max(wecare_nfe_variants_inv.df$af_inv) # 0.5

max(apply(wecare_nfe_genotypes.mx,1,sum,na.rm=TRUE)) # 1,351
max(apply(wecare_nfe_genotypes_inv.mx,1,sum,na.rm=TRUE)) # 673

# clean-up
rm(var)

```

# prepare_data_for_analysis

No missed data in eigenvectors and outcomes  
Impute missed data for genotypes  

```{r prepare_data_for_analysis}

# Eigenvectors
E <- as.matrix(wecare_nfe_phenotypes.df[,c("eig1_nfe","eig2_nfe","eig3_nfe")])
E[1:5,]
sum(is.na(E)) # 0

# Outcomes
Y <- ifelse(wecare_nfe_phenotypes.df$cc==-1, 0, 1)
sum(is.na(Y)) # 0
sum(Y) # 478
sum(!Y) # 198

# Genotypes: transpose non-aggregated variants
G <- t(wecare_nfe_genotypes_inv.mx)

# Genotypes: impute missed data
#G <- apply(G, 2, function(v) { ifelse(is.na(v), mean(v, na.rm=T), v) })
for(col in 1:ncol(G)){
  genotypes <- G[,col]
  mean_genotype <- mean(genotypes, na.rm=TRUE)
  mean_genotype -> genotypes[is.na(genotypes)]
  genotypes -> G[,col]
  rm(genotypes, mean_genotype)
}

# Check data
dim(G)
G[1:5,1:5]
sum(is.na(G)) # 0

# Clean-up
rm(col)

```

# prepare_file_and_matrix_for_results

```{r prepare_file_and_matrix_for_results}

# Prepare header
header <- c("variant_id", "ac_inv", "p_glm_wald", "p_glm_anova_lrt", "p_glm_lrt")

# Write header to output matrix 
# (do not use data frame here: it will play with rbind!)
result.mx <- matrix(ncol=length(header), nrow=0) 
colnames(result.mx) <- header

# Write header to output file
results_file="r08a_wecare_vs_nfe_all_vars_glm_3ev_nfe_Apr2017.txt"
write(paste(header, sep="", collapse="\t"), 
      file=paste(results_folder, results_file, sep="/"))

# Clean-up
rm(header)

```

# logit_regression

Reccurent warnings:  
Warning: glm.fit: fitted probabilities numerically 0 or 1 occurred  

```{r logit_regression, warning=FALSE}

#G <- G[,1:100]

# For each variant
for(var in 1:ncol(G)){

  # Get vector of genotypes  
  X <- G[,var]

  var_id <- as.character(wecare_nfe_variants_inv.df[var,"SplitVarID"])
  ac_inv <- wecare_nfe_variants_inv.df[var,"ac_inv"]

  # --- Calculate the regressions --- #
  
  regE <- glm(Y ~ E, family=binomial) # Null model
  regXE <- glm(Y ~ X + E, family=binomial) # Complete model

  # Wald (?) test
  p_glm_wald <- summary(regXE)$coef["X", "Pr(>|z|)"]
  
  # Anova-LRT
  anova_lrt <- anova(regXE, test="LRT")
  p_glm_anova_lrt <- anova_lrt["X", "Pr(>Chi)"]
  
  # LRT
  chi_stat_lrt = 2*(logLik(regXE) - logLik(regE))
  p_glm_lrt = 1-pchisq(chi_stat_lrt, df=1) # df??
  
  # Clean-up
  rm(regE, regXE, anova_lrt, chi_stat_lrt)
  
  # --- Record results (p being converted to char) --- #
  
  # Add result to matrix
  result.mx <- rbind(result.mx, c(var_id, ac_inv, p_glm_wald, p_glm_anova_lrt, p_glm_lrt))
  
  # Add result to file
  write(paste(c(var_id, ac_inv, p_glm_wald, p_glm_anova_lrt, p_glm_lrt), sep="", collapse="\t"), 
        file=paste(results_folder, results_file, sep="/"), append = TRUE)

  # Clean-up
  rm(X, p_glm_wald, p_glm_anova_lrt, p_glm_lrt, var_id)
  
} # next variant

# Clean-up
rm(var, G, Y, E, results_file)

```

# explore_result

```{r explore_result}

# Convert to data.frame and to numeric
result.df <- as.data.frame(result.mx)

result.df$variant_id <- as.vector(result.df$variant_id)

result.df$p_glm_wald <- as.numeric(as.vector(result.df$p_glm_wald))
result.df$p_glm_anova_lrt <- as.numeric(as.vector(result.df$p_glm_anova_lrt))
result.df$p_glm_lrt <- as.numeric(as.vector(result.df$p_glm_lrt))

str(result.df)

# Sort result
result.df <- result.df[order(result.df$p_glm_lrt),]
result.df[1:10,]

# Look at correlation of diferent p-estimates
# Anova does not correlate with anything
# glm(wald?) more or less correlate with lrt

plot(p_glm_wald~p_glm_anova_lrt, data=result.df)
plot(p_glm_wald~p_glm_lrt, data=result.df)
plot(p_glm_lrt~p_glm_anova_lrt, data=result.df)

# Look at histograms of p-values
hist(result.df$p_glm_wald)
hist(result.df$p_glm_anova_lrt)
hist(result.df$p_glm_lrt)

sum(result.df$p_glm_wald == 0)
sum(result.df$p_glm_anova_lrt == 0)
sum(result.df$p_glm_lrt == 0)

sum(result.df$p_glm_wald == 1)
sum(result.df$p_glm_anova_lrt == 1)
sum(result.df$p_glm_lrt == 1)

# Load function for nice qq-plots
source("f03_qqunif_plot.R")

p_wald <- result.df$p_glm_wald[!result.df$p_glm_wald == 0]
p_anova_lrt <- result.df$p_glm_anova_lrt[!result.df$p_glm_anova_lrt == 0]
p_lrt <- result.df$p_glm_lrt[!result.df$p_glm_lrt == 0]

qqunif.plot(p_wald, main="WECARE vs NFE\n(274,037x676, all vars, 3 nfe eigenvectors, glm wald)")
qqunif.plot(p_anova_lrt, main="WECARE vs NFE\n(274,037x676, all vars, 3 nfe eigenvectors, glm anova-lrt)")
qqunif.plot(p_lrt, main="WECARE vs NFE\n(274,037x676, all vars, 3 nfe eigenvectors, glm lrt)")

# Clean-up
rm(p_wald, p_anova_lrt, p_lrt, result.mx)

```

# data_summary

```{r data_summary}

dim(wecare_nfe_genotypes.mx)
class(wecare_nfe_genotypes.mx)
wecare_nfe_genotypes.mx[1:5,1:5]

dim(wecare_nfe_genotypes_inv.mx)
class(wecare_nfe_genotypes_inv.mx)
wecare_nfe_genotypes_inv.mx[1:5,1:5]

dim(wecare_nfe_kgen.df)
colnames(wecare_nfe_kgen.df)
wecare_nfe_kgen.df[1:5,1:5]

dim(wecare_nfe_exac.df)
colnames(wecare_nfe_exac.df)
wecare_nfe_exac.df[1:5,1:5]

dim(wecare_nfe_variants.df)
str(wecare_nfe_variants.df)
wecare_nfe_variants.df[1:5,1:5]

dim(wecare_nfe_variants_inv.df)
str(wecare_nfe_variants_inv.df)
wecare_nfe_variants_inv.df[1:5,1:5]

dim(wecare_nfe_phenotypes.df)
str(wecare_nfe_phenotypes.df)
wecare_nfe_phenotypes.df[1:5,1:5]

dim(result.df)
str(result.df)
result.df[1:5,]

# Check consistency of rownames and colnames
sum(colnames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_phenotypes.df))

sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_kgen.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_exac.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants.df))
sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_variants_inv.df))

sum(rownames(wecare_nfe_genotypes.mx) != rownames(wecare_nfe_genotypes_inv.mx))
sum(colnames(wecare_nfe_genotypes.mx) != colnames(wecare_nfe_genotypes_inv.mx))

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "s08a_wecare_vs_nfe_all_vars_glm_3ev_nfe_Apr2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
