---
title: "matched_analysis_wecare_only"
output: html_document
---

started: Alexey Larionov, 27Feb2017  
last updated: Alexey Larionov, 26Apr2017

# Summary

Matched analysis with clogit on SKAT-style weighted and aggregated data  

- Read and check the data  
- Perform matched analysis on weighted and aggregated data  
- Compare matched analysis with un-matched glm (Firth?) based on David Conti e-mail of 18Nov2016  
- Make QQ-plots  
- Write results into text files  

# start_section

```{r start_section}

# Start time
Sys.time()

# Folders
#/analysis/mtgroup_share
#/scratch/medgen
setwd("/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts/wecare_only")
source_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/source_data"
interim_data_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts/wecare_only/interim_data"
results_folder <- "/analysis/mtgroup_share/scripts/wecare_stat_04.17/scripts/wecare_only/results"

# Load libraries and functions
library(survival) # for clogit (matched analysis)
source("f02_tryCatchAdv.R") # advanced error handling
source("f03_qqunif_plot.R") # for nice qq-plots

```

# load_data

```{r load_data}

load(paste(interim_data_folder, "s10_invert_impute_weight_aggregate_wecare_only.RData", sep="/"))

```

# check_data

```{r check_data}

ls()

dim(genotypes.mx)
class(genotypes.mx)
genotypes.mx[1:5,1:5]

dim(genotypes_inv_imp_wt.mx)
class(genotypes_inv_imp_wt.mx)
genotypes_inv_imp_wt.mx[1:5,1:5]

dim(genes_aggr_data.mx)
class(genes_aggr_data.mx)
genes_aggr_data.mx[1:5,1:5]

dim(genes_aggr_info.df)
str(genes_aggr_info.df)
genes_aggr_info.df[1:5,1:5]

dim(kgen.df)
colnames(kgen.df)
kgen.df[1:5,1:5]

dim(exac.df)
colnames(exac.df)
exac.df[1:5,1:5]

dim(variants.df)
str(variants.df)
variants.df[1:5,1:5]

dim(phenotypes.df)
str(phenotypes.df)
phenotypes.df[1:5,1:5]

# Check consistency of rownames and colnames
sum(colnames(genotypes.mx) != rownames(phenotypes.df))
sum(colnames(genes_aggr_data.mx) != rownames(phenotypes.df))

sum(rownames(genes_aggr_info.df) != rownames(genes_aggr_data.mx))

sum(rownames(genotypes.mx) != rownames(genotypes_inv_imp_wt.mx))
sum(rownames(genotypes.mx) != rownames(kgen.df))
sum(rownames(genotypes.mx) != rownames(exac.df))
sum(rownames(genotypes.mx) != rownames(variants.df))

```

# prepare_data_for_clogit_and_glm

Verifying that there is No missed data   

```{r prepare_data_for_analysis}

# Eigenvectors
E <- as.matrix(phenotypes.df[,c("eig1", "eig2")])
sum(is.na(E)) # 0

# Pairs IDs (for clogit)
P <- phenotypes.df$setno
sum(is.na(P)) # 0

# Matching information (for glm)
M <- as.matrix(phenotypes.df[,c("age_dx", "rstime")])
sum(is.na(M)) # 0

# Make a variable for hormone treatment.
hormone <- phenotypes.df$hormone

# Make variable for cytotoxic treatment
chemo <- phenotypes.df$chemo_cat != "no"

# Make joined treatment variable
treatment <- as.integer(chemo | hormone)
sum(is.na(treatment))

# Number of pregnancies
num_preg <- as.integer(phenotypes.df$num_preg)
sum(is.na(num_preg))

# Bind clinical covariates together
C <- cbind(treatment, num_preg)

# Outcomes
Y <- phenotypes.df$cc
sum(is.na(Y))

# Clean-up
rm(chemo, hormone, treatment, num_preg)

```

# run_clogit

Use advanced error handling (see tryCatchAdv function)  
~3-5 min  

```{r run_clogit}

# Get list of genes (get rid of NAs !)
genes <- unique(as.character(variants.df$SYMBOL))
length(genes)
sum(is.na(genes))

# Prepare output matrix
header <- c("gene", "clogit_mode", "p_clogit_lrt", "p_clogit_anova", "p_clogit_wald", 
            "std_clogit_status", "std_clogit_message", "std_clogit_null_status", "std_clogit_null_message", 
            "appr_clogit_status", "appr_clogit_message", "appr_clogit_null_status", "appr_clogit_null_message")

clogit.mx <- matrix(nrow=0, ncol=length(header))
header -> colnames(clogit.mx)  

#genes <- c("ATM", "CASP9", "ADH1B", "SLC27A3", "SLC9A3R1", "CCDC27", "TTLL10")

# For each gene
for(gene in genes){
  
  #gene="CCDC27"
  #gene="ATM"
  
  # Get explanatory vector  
  X <- genes_aggr_data.mx[gene,]
  
  # Initialise output variables
  clogit_mode <- "standard"
  clogit_reg <- NA
  clogit_reg_null <- NA
  p_clogit_lrt <- NA
  p_clogit_anova <- NA
  p_clogit_wald <- NA
  std_clogit_status <- NA
  std_clogit_message <- NA
  std_clogit_null_status <- NA
  std_clogit_null_message <- NA
  appr_clogit_status <- NA
  appr_clogit_message <- NA
  appr_clogit_null_status <- NA
  appr_clogit_null_message <- NA
  
  # Try clogit using advanced catching of errors/warnings
  # (see details within the tryCatchAdv() function)
  clogit_reg <- tryCatchAdv(clogit( Y ~ X + E + C + strata(P)))
  clogit_reg_null <- tryCatchAdv(clogit( Y ~ E + C + strata(P)))
  
  # Get status (success or error/warning) and error/warning messages (if any)
  std_clogit_status <- clogit_reg$status
  std_clogit_message <- clogit_reg$message
  std_clogit_null_status <- clogit_reg_null$status
  std_clogit_null_message <- clogit_reg_null$message
  
  # If any clogit model failed in default ("standard") mode e.g.
  # http://r.789695.n4.nabble.com/Error-when-running-Conditional-Logit-Model-td948556.html 
  if( std_clogit_status == "error" | std_clogit_null_status == "error"){

    # Try approximate clogit
    # http://stats.stackexchange.com/questions/109222/conditional-logistic-regression-model-does-not-converge-but-logistic-regression
    clogit_mode <- "approximate"
    
    clogit_reg <- tryCatchAdv(clogit( Y ~ X + E + C + strata(P), method="approximate"))
    clogit_reg_null <- tryCatchAdv(clogit( Y ~ E + C + strata(P), method="approximate"))

    # Get status and messages    
    appr_clogit_status <- clogit_reg$status
    appr_clogit_message <- clogit_reg$message
    appr_clogit_null_status <- clogit_reg_null$status
    appr_clogit_null_message <- clogit_reg_null$message

  }

  # If both standard and approximate attempts failed
  if((std_clogit_status == "error" | std_clogit_null_status == "error") & 
     (appr_clogit_status == "error" | appr_clogit_null_status == "error")){
    
    # Set eror status, do not calculate p-estimates
    clogit_mode <- "error"
    
  # Calculate p-estimates
  } else {
    
    # Extract values from tryCatchAdv objects
    clogit_reg <- clogit_reg$value
    clogit_reg_null <- clogit_reg_null$value
    
    # Calculate LRT p-estimate
    chi_stat_lrt = 2*(logLik(clogit_reg) - logLik(clogit_reg_null))
    p_clogit_lrt = 1 - pchisq(chi_stat_lrt, df=1)
    
    # Calculate LRT p-estimate using anova()
    clogit_anova <- anova(clogit_reg, clogit_reg_null)
    p_clogit_anova <- clogit_anova[2,"P(>|Chi|)"]
    
    # Get Wald p-estimate
    p_clogit_wald <- summary(clogit_reg)$coef["X", "Pr(>|z|)"]
  
    # Clean-up
    rm(chi_stat_lrt, clogit_anova)

  }

  # Extract text from message objects (if any)
  if(!is.na(std_clogit_status) &
     (std_clogit_status == "warning" | 
      std_clogit_status == "error")){
    std_clogit_message <- std_clogit_message$message}
  
  if(!is.na(std_clogit_null_status) & 
     (std_clogit_null_status == "warning" |
      std_clogit_null_status == "error")){
    std_clogit_null_message <- std_clogit_null_message$message}

  if(!is.na(appr_clogit_status) &
     (appr_clogit_status == "warning" | 
      appr_clogit_status == "error")){
    appr_clogit_message <- appr_clogit_message$message}
  
  if(!is.na(appr_clogit_null_status) &
     (appr_clogit_null_status == "warning" |
      appr_clogit_null_status == "error")){
    appr_clogit_null_message <- appr_clogit_null_message$message}
  
  # Compile result
  result <- c(gene, clogit_mode, p_clogit_lrt, p_clogit_anova, p_clogit_wald, 
            std_clogit_status, std_clogit_message, std_clogit_null_status, std_clogit_null_message, 
            appr_clogit_status, appr_clogit_message, appr_clogit_null_status, appr_clogit_null_message)
  
  # Add result to output table
  clogit.mx <- rbind(clogit.mx, result)

  # Clean-up
  rm(X, clogit_mode, clogit_reg, clogit_reg_null, p_clogit_lrt, p_clogit_anova, p_clogit_wald, 
     std_clogit_status, std_clogit_message, std_clogit_null_status, std_clogit_null_message, 
     appr_clogit_status, appr_clogit_message, appr_clogit_null_status, appr_clogit_null_message,
     result)

} # Next gene

# Add rownames
genes -> rownames(clogit.mx)

# Check the result
dim(clogit.mx)

# Convert to data frame and update data types
clogit_result.df <- as.data.frame(clogit.mx)
clogit_result.df$gene <- as.vector(clogit_result.df$gene)
clogit_result.df$p_clogit_lrt <- as.numeric(as.vector(clogit_result.df$p_clogit_lrt))
clogit_result.df$p_clogit_anova <- as.numeric(as.vector(clogit_result.df$p_clogit_anova))
clogit_result.df$p_clogit_wald <- as.numeric(as.vector(clogit_result.df$p_clogit_wald))
str(clogit_result.df)

# Explore results
summary(clogit_result.df$clogit_mode)

summary(clogit_result.df$p_clogit_lrt)
summary(clogit_result.df$p_clogit_anova)
sum(clogit_result.df$p_clogit_anova == 0, na.rm=TRUE)
summary(clogit_result.df$p_clogit_wald)

summary(clogit_result.df$std_clogit_status)
summary(clogit_result.df$std_clogit_message[clogit_result.df$std_clogit_status=="error", drop=TRUE])
summary(clogit_result.df$std_clogit_message[clogit_result.df$std_clogit_status=="warning", drop=TRUE])
# 456 genes where "Loglik converged before variable  1 ; beta may be infinite." - are these very good genes?

summary(clogit_result.df$std_clogit_null_status)
summary(clogit_result.df$std_clogit_null_message)

summary(clogit_result.df$appr_clogit_status)
summary(clogit_result.df$appr_clogit_message[clogit_result.df$appr_clogit_status=="error", drop=TRUE])
summary(clogit_result.df$appr_clogit_message[clogit_result.df$appr_clogit_status=="warning", drop=TRUE])
# 624 genes "Loglik converged before variable ... ; beta may be infinite." - are these very good genes?

summary(clogit_result.df$appr_clogit_null_status)
summary(clogit_result.df$appr_clogit_null_message)

# Write result to file
output_file <- paste(results_folder, "r11_matched_analysis_wecare_only.txt", sep="/")
write.table(clogit_result.df, file=output_file, quote=FALSE, sep="\t")

# Clean-up
rm(P, output_file, header, clogit.mx)

```

# compare_clogit_p_estimates

```{r compare_clogit_p_estimates}

p_clogit_lrt <- clogit_result.df$p_clogit_lrt
sum(!is.na(p_clogit_lrt))

p_clogit_lrt_standard <- p_clogit_lrt[clogit_result.df$clogit_mode=="standard"]
sum(!is.na(p_clogit_lrt_standard))

p_clogit_anova <- clogit_result.df$p_clogit_anova
sum(!is.na(p_clogit_anova))

p_clogit_anova_standard <- p_clogit_anova[clogit_result.df$clogit_mode=="standard"]
sum(!is.na(p_clogit_anova_standard))

p_clogit_wald <- clogit_result.df$p_clogit_wald
sum(!is.na(p_clogit_wald))

p_clogit_wald_standard <- p_clogit_wald[clogit_result.df$clogit_mode=="standard"]
sum(!is.na(p_clogit_wald_standard))

main=paste("clogit ANOVA vs LRT\nall",sum(!is.na(p_clogit_lrt)),"genes")
plot(p_clogit_lrt,p_clogit_anova, main=main)

main=paste("clogit ANOVA vs LRT (log10 scale)\nall",sum(!is.na(p_clogit_lrt)),"genes")
plot(log(p_clogit_lrt, 10),log(p_clogit_anova, 10), main=main)

main=paste("clogit ANOVA vs LRT\n",
           sum(!is.na(p_clogit_lrt_standard)),"genes in standard clogit mode")
plot(p_clogit_lrt_standard,p_clogit_anova_standard, main=main)

main=paste("clogit ANOVA vs LRT (log10 scale)\n",
           sum(!is.na(p_clogit_lrt_standard)),"genes in standard clogit mode")
plot(log(p_clogit_lrt_standard, 10),log(p_clogit_anova_standard, 10), main=main)

main=paste("clogit Wald vs LRT\nall",sum(!is.na(p_clogit_wald)),"genes")
plot(p_clogit_lrt,p_clogit_wald, main=main)

main=paste("clogit Wald vs LRT (log10 scale)\nall",sum(!is.na(p_clogit_wald)),"genes")
plot(log(p_clogit_lrt, 10),log(p_clogit_wald, 10), main=main)

main=paste("clogit Wald vs LRT\n",sum(!is.na(p_clogit_wald_standard)),"genes in standard clogit mode")
plot(p_clogit_lrt_standard,p_clogit_wald_standard, main=main)

main=paste("clogit Wald vs LRT (log10 scale)\n",
           sum(!is.na(p_clogit_wald_standard)),"genes in standard clogit mode")
plot(log(p_clogit_lrt_standard, 10),log(p_clogit_wald_standard,10), main=main)

rm(p_clogit_lrt, p_clogit_lrt_standard,
   p_clogit_anova, p_clogit_anova_standard,
   p_clogit_wald, p_clogit_wald_standard)

```

# clogit_qq_plots

```{r clogit_qq_plots}

# lrt
p_clogit_lrt <- clogit_result.df$p_clogit_lrt
lrt_na_count <- sum(is.na(p_clogit_lrt))
lrt_na_count
lrt_zero_count <- sum(p_clogit_lrt==0, na.rm=TRUE)
lrt_zero_count
p_clogit_lrt <- p_clogit_lrt[!is.na(p_clogit_lrt)]
p_clogit_lrt <- p_clogit_lrt[p_clogit_lrt != 0]
main <- paste("CBC vs UBC: matched, clogit lrt\n", 
              length(p_clogit_lrt),"genes (",
              lrt_na_count, "NAs and",
              lrt_zero_count, "zeroes omitted)")
qqunif.plot(p_clogit_lrt, main=main)

# anova
p_clogit_anova <- clogit_result.df$p_clogit_anova
anova_na_count <- sum(is.na(p_clogit_anova))
anova_na_count
anova_zero_count <- sum(p_clogit_anova==0, na.rm=TRUE)
anova_zero_count
p_clogit_anova <- p_clogit_anova[!is.na(p_clogit_anova)]
p_clogit_anova <- p_clogit_anova[p_clogit_anova != 0]
main <- paste("CBC vs UBC: matched, clogit anova\n", 
              length(p_clogit_anova),"genes (",
              anova_na_count, "NAs and",
              anova_zero_count, "zeroes omitted)")
qqunif.plot(p_clogit_anova, main=main)

# wald
p_clogit_wald <- clogit_result.df$p_clogit_wald
wald_na_count <- sum(is.na(p_clogit_wald))
wald_na_count
wald_zero_count <- sum(p_clogit_wald==0, na.rm=TRUE)
wald_zero_count
p_clogit_wald <- p_clogit_wald[!is.na(p_clogit_wald)]
p_clogit_wald <- p_clogit_wald[p_clogit_wald != 0]
main <- paste("CBC vs UBC: matched, clogit wald\n", 
              length(p_clogit_wald),"genes (",
              wald_na_count, "NAs and",
              wald_zero_count, "zeroes omitted)")
qqunif.plot(p_clogit_wald, main=main)

# Clean-up
rm(lrt_na_count, lrt_zero_count, p_clogit_lrt, p_clogit_lrt_standard,
   anova_na_count, anova_zero_count, p_clogit_anova, p_clogit_anova_standard,
   wald_na_count, wald_zero_count, p_clogit_wald, p_clogit_wald_standard)

```






# explore_results

```{r explore_results}

# Read results
results.df <- read.table(paste(results_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.txt", sep="/"), header = TRUE)
dim(results.df)
str(results.df)

summary(as.factor(results.df$multilaalelic_varaints))
# FALSE  TRUE 
#  8463   303 

summary(as.factor(results.df$inverted_afs))
# FALSE  TRUE 
#  8618   148 

summary(as.factor(results.df$clogit_mode))
#  approximate  failed_anova failed_clogit      standard 
#          601            19            45          8,101 

summary(as.factor(results.df$glm_mode)) 
# standard: 8,766

# ------------ Compare P values ------------ #
mean_p_values <- c(
  mean(results.df$p_clogit, na.rm=TRUE),
  mean(results.df$p_clogit_anova, na.rm=TRUE),
  mean(results.df$p_glm),
  mean(results.df$p_glm_lrt))
names(mean_p_values) <- c("clogit", "clogit_anova", "glm", "glm_lrt")
barplot(mean_p_values)

colours <- as.vector(results.df$clogit_mode)
"black" -> colours[colours == "standard"]
"red" -> colours[colours == "approximate"]
"green" -> colours[colours == "failed_clogit"]
"green" -> colours[colours == "failed_anova"]

plot(p_glm_lrt ~ p_clogit_anova, data=results.df,
     col=colours, main="glm_lrt vs clogit_anova")
abline(a=0, b=1, col="blue", lwd=3)
legend("bottomright", inset=.05, 
       title="clogit mode:",
       c("standard", "approximate"), 
       fill=c("black", "red"),
       horiz = TRUE)

plot(p_clogit ~ p_clogit_anova, data=results.df, 
     col=colours, main="p_clogit vs clogit_anova")
abline(a=0, b=1, col="blue", lwd=3)
legend("bottomright", inset=.05, 
       title="clogit mode:",
       c("standard", "approximate"), 
       fill=c("black", "red"),
       horiz = TRUE)

plot(p_glm ~ p_glm_lrt, data=results.df, 
     col=colours, main="p_glm vs p_glm_lrt")
abline(a=0, b=1, col="blue", lwd=3)
legend("bottomright", inset=.05, 
       title="clogit mode:",
       c("standard", "approximate", "failed"), 
       fill=c("black", "red", "green"),
       horiz = TRUE)

# ---------------- qq plots ---------------- #

# http://genome.sph.umich.edu/wiki/Code_Sample:_Generating_QQ_Plots_in_R
# http://www.gettinggeneticsdone.com/2014/05/qqman-r-package-for-qq-and-manhattan-plots-for-gwas-results.html 

# It is belived that under null hypothesis the distribution of P-values shoulod be uniform
# e.g. http://physics.rockefeller.edu/luc/technical_reports/cdf8023_facts_about_p_values.pdf
# Hopefully, this holds for the GWAs-WES type of data

obs.pvalues <- as.numeric(results.df$p_glm_lrt)
exp.pvalues<-(rank(obs.pvalues, ties.method="first") + 0.5)/(length(obs.pvalues)+1)# # note the flat expectation
plot(-log10(exp.pvalues), -log10(obs.pvalues), 
     xlim = c(0,4.5), ylim = c(0,4.5),
     main="CBC vs UBC: p_glm_lrt")
abline(0,1,col="red")

qqunif.plot(obs.pvalues, main="CBC vs UBC: p_glm_lrt")

obs.pvalues <- as.numeric(results.df$p_clogit_anova[!is.na(results.df$p_clogit_anova)])
exp.pvalues<-(rank(obs.pvalues, ties.method="first") + 0.5)/(length(obs.pvalues)+1)
plot(-log10(exp.pvalues), -log10(obs.pvalues), 
     xlim = c(0,4.5), ylim = c(0,4.5),
     main="CBC vs UBC: p_clogit_anova")
abline(0,1,col="red")

qqunif.plot(obs.pvalues, main="CBC vs UBC: p_clogit_anova")

# --- Sort --- #

results.df <- results.df[order(as.numeric((results.df$p_glm_lrt))), ] 

# Save sorted
write.table(results.df, 
            file=paste(results_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.txt", sep="/"),
            row.names = FALSE, quote=FALSE, sep="\t")

```

# save_data

```{r save_data}

save.image(paste(interim_data_folder, "s08a_3_matched_analysis_wecare_only_Mar2017.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```
